---
title: "Differential Gene Expression Analysis"
format:
  html:
    theme: flatly
    toc: true
    toc-location: left
    toc-depth: 4
execute:
  echo: false
  warning: false
  message: true
params:
  # counts_file: ".test/counts_matrix.tsv"
  # samplesheet: ".test/samples.tsv"
  counts_file: "~/Desktop/250427/counts_matrix.tsv"
  samplesheet: "~/Desktop/250427/samples.tsv"
  useERCC: TRUE
  group1: "KOS"
  group2: "Uninf"
  sample_column: "sampleName"
  group_column: "groupName"
  batch: TRUE
  batch_column: "batch"
  design_formula: "~ groupName"
  outdir: "~/Desktop/250427/results_250427"
  # normalized_counts_file: "normalized_counts.tsv"
  # deg_results_file: "deg_results.tsv"
  log2FC_threshold: 1
  pvalue_threshold: 0.05
  fdr_threshold: 0.05
  edgeR_cpm_cutoff: 1
  edgeR_cpm_group_fraction: 0.5
  DESeq2_low_count_cutoff: 2
  DESeq2_low_count_group_fraction: 0.5
  prerank_choice: "edgeR" # choices: "edgeR", "DESeq2", "combined"
  host: "Hs" # choices: "Hs", "Mm"
  genes_selection: "host" # choices: "host", "virus", "both" # filter out genes other than this selection
---

# 🔬 Differential Expression Analysis

This report performs standard DGE analysis using input count data, metadata, and user-defined contrasts.

---
```{r library}
#| code-fold: true
#| code-summary: "📦 Load Required Packages"
suppressPackageStartupMessages({
  library(limma)
  library(edgeR)
  library(DESeq2)
  library(tibble)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(plotly)
  library(EnhancedVolcano)
  library(knitr)
  library(kableExtra)
  library(ggplotify)
  library(ggpubr)
  library(ggrepel)
  library(openxlsx)
  library(glue)
  library(DT)
  library(ComplexHeatmap)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
  library(msigdbr)
  library(UpSetR)
  library(ggplot2)
  library(plotly)
  library(openxlsx)
  library(yaml)
  library(fs)
  library(VennDiagram)
  library(grid)
})

exit_gracefully <- function() {
  message("Exiting early: Debugging script intentionally halted.")
  knitr::knit_exit()
}
# Paths
output_base <- params$outdir
counts_dir <- file.path(output_base, "counts")
edger_deg_dir <- file.path(output_base, "edgeR_deg")
deseq2_deg_dir <- file.path(output_base, "DESeq2_deg")
# gsea_dir <- file.path(output_base, "gsea")

# Create folders
fs::dir_create(c(output_base,counts_dir, edger_deg_dir, deseq2_deg_dir))

# write out params
# Convert Quarto parameters (params) to a named list
yaml_params <- as.list(params)

# Write to YAML file
yaml::write_yaml(yaml_params, file.path(output_base, "params.yaml"))
```

## Reading inputs

```{r readinput}
#| code-fold: true
#| code-summary: "📥 Read Count Matrix and Sample Sheet"
#| message: true
#| warning: false

print(glue::glue("📥 Reading input files: "))
print(glue::glue(params$counts_file, " and "))
print(glue::glue(params$samplesheet))

counts <- read.table(params$counts_file,
                     check.names = FALSE,  # prevent automatic conversion of hyphen to dot
                     row.names = 1, 
                     sep = "\t", 
                     header = TRUE)

original_counts <- counts
# Report original dimensions
print(glue::glue("📊 Counts matrix loaded: ", nrow(counts), " genes (rows) × ", ncol(counts), " samples (columns)"))

# Replace hyphens with underscores in column names
colnames(counts) <- gsub("-", "_", colnames(counts))

# Read the sample sheet
samples <- read.table(params$samplesheet, check.names = FALSE, sep = "\t", header = TRUE)
keep_cols <- c(params$sample_column, params$group_column)
if (isTRUE(params$batch)) {
  keep_cols <- c(keep_cols, params$batch_column)
}
# Find which of the required columns are actually present
present_cols <- intersect(keep_cols, colnames(samples))

# Check: all required columns must be present
if (length(present_cols) != length(keep_cols)) {
  missing_cols <- setdiff(keep_cols, present_cols)
  stop("❌ The following required columns are missing from 'samples': ", paste(missing_cols, collapse = ", "))
}

# Subset to the required columns
samples <- samples[, present_cols, drop = FALSE]

# Cleanup column values
samples[[params$sample_column]] <- gsub("-", "_", samples[[params$sample_column]])
samples[[params$group_column]]  <- gsub("-", "_", samples[[params$group_column]])

```

## Filtering by DEG groups

```{r filtercounts}
#| code-fold: true
#| code-summary: "🎯 Filter samples and counts matrix for selected groups"
#| message: true
#| warning: false

# Get groups from params
g1 <- params$group1
g2 <- params$group2

# Define color palette based on actual group names
group_colors <- setNames(c("#0072B2", "#D55E00"), c(g1, g2))

print(glue::glue("📥 Selecting samples in groups: "))
print(glue::glue(g1, " and "))
print(glue::glue(g2))

# Filter the samplesheet to only include samples from group1 and group2
selected_samples <- samples[samples[[params$group_column]] %in% c(g1, g2), ]

# Get the sample IDs (assuming there's a 'sample' column)
sample_ids <- selected_samples$sample

# Now subset the counts matrix
# Ensure sample columns in counts match sample IDs
filtered_counts <- counts[, colnames(counts) %in% sample_ids]
# Report dimensions
print(glue::glue("📦 Filtered counts matrix: ", nrow(filtered_counts), " genes × ", ncol(filtered_counts), " samples"))
```

## Filtering by genes selection

```{r genes_selection}
#| code-fold: true
#| code-summary: "🔍 Filter by genes_selection"
#| message: true
#| warning: false

print(glue::glue("Genes selection: ", params$genes_selection))
print(glue::glue("useERCC: ", params$useERCC))

# Filter out genes based on the genes_selection
if (params$genes_selection == "host") {
  if (isTRUE(params$useERCC)) {
    # Select only host and ERCC genes
    if (params$host == "Hs") {
      filtered_counts <- filtered_counts[grepl("^ENSG|^ERCC", rownames(filtered_counts)), ]
    } else if (params$host == "Mm") {
      filtered_counts <- filtered_counts[grepl("^ENSMUSG|^ERCC", rownames(filtered_counts)), ]
    }
  } else {
    # Select only host genes
    if (params$host == "Hs") {
      filtered_counts <- filtered_counts[grepl("^ENSG", rownames(filtered_counts)), ]
    } else if (params$host == "Mm") {
      filtered_counts <- filtered_counts[grepl("^ENSMUSG", rownames(filtered_counts)), ]
    }
  }
  nercc_genes <- nrow(filtered_counts[grepl("^ERCC", rownames(filtered_counts)), ])
  print(glue::glue("🦠 Host-only genes selected: ", nrow(filtered_counts), " genes remaining (including ", nercc_genes, " ERCC genes)"))
} else if (params$genes_selection == "virus") {
  if (isTRUE(params$useERCC)) {
    # Filter out host genes .. keep ERCC
    filtered_counts <- filtered_counts[!grepl("^ENS", rownames(filtered_counts)), ]
  } else {
    # Filter out host genes and ERCC genes
    filtered_counts <- filtered_counts[!grepl("^ERCC|^ENS", rownames(filtered_counts)), ]
  }
  nercc_genes <- nrow(filtered_counts[grepl("^ERCC", rownames(filtered_counts)), ])
  print(glue::glue("🦠 Virus-only genes selected: ", nrow(filtered_counts), " genes remaining (including ", nercc_genes, " ERCC genes)"))
} else if (params$genes_selection == "both") {
  if (!isTRUE(params$useERCC)) {
    # Keep both host and virus genes .. remove ERCC
    filtered_counts <- filtered_counts[grepl("^ERCC", rownames(filtered_counts)), ]
  }
  nercc_genes <- nrow(filtered_counts[grepl("^ERCC", rownames(filtered_counts)), ])
  print(glue::glue("🦠 Both host and virus genes included: ", nrow(filtered_counts), " genes remaining (including ", nercc_genes, " ERCC genes)"))
} else {
  stop("❌ Invalid mode selected. Choose 'host', 'virus', or 'both'.")
}
# Report dimensions
print(glue::glue("📦 Filtered counts matrix: ", nrow(filtered_counts), " genes × ", ncol(filtered_counts), " samples"))

counts <- filtered_counts
is_ercc <- grepl("^ERCC", rownames(counts))
is_endogenous <- !is_ercc
is_nonzero_ercc <- is_ercc & (rowSums(counts) > 0)
samples <- selected_samples
samples$groupName <- relevel(factor(samples[[params$group_column]], levels = c(g1, g2)),ref = g2)
#exit_gracefully()
if (isTRUE(params$useERCC)) {
  print(glue::glue("❌ Warning: ",sum(is_ercc) - sum(is_nonzero_ercc), " ERCC genes have ZERO counts in all samples!."))

    # Check if any ERCC genes have zero counts in all samples
  if (sum(is_nonzero_ercc) == 0) {
    stop("❌ All ERCC genes have zero counts in all samples. Please check your data.")
  }
  print(glue::glue("✅ ", sum(is_nonzero_ercc), " ERCC genes have non-zero counts in at least one sample."))
} else {
  # Check if any ERCC genes have zero counts in all samples
  if (sum(is_ercc) > 0) {
    print(glue::glue("❌ ",sum(is_ercc)," ERCC genes are present but will be removed before analysis."))
  }
  counts <- counts[!is_ercc, ]
  # update
  is_ercc <- grepl("^ERCC", rownames(counts))
  is_endogenous <- !is_ercc
  is_nonzero_ercc <- is_ercc & (rowSums(counts) > 0)
}
```
## Validating group information

```{r validateinput}
#| code-fold: true
#| code-summary: "✅ Validate Contrast Groups"
#| message: true
#| warning: false
# Check unique group levels in the samplesheet
valid_groups <- unique(samples[[params$group_column]])

# Find missing groups
missing_groups <- setdiff(c(g1, g2), valid_groups)

if (length(missing_groups) > 0) {
  stop(paste0("❌ Group(s) from params not found in samples[[params$group_column]]\n",
       paste(missing_groups, collapse = ", ")))
} else {
  print(glue::glue("✅ Both group1(",g1,") and group2(",g2,") are present in samplesheet."))
  
  # Check if each group has at least 2 samples
  group_counts <- table(samples$groupName)
  too_few <- c()
  if (group_counts[g1] < 2) too_few <- c(too_few, g1)
  if (group_counts[g2] < 2) too_few <- c(too_few, g2)
  
  if (length(too_few) > 0) {
    stop("❌ The following group(s) have fewer than 2 samples:\n",
         paste(too_few, collapse = ", "))
  } else {
    print(glue::glue("✅ Each group has at least 2 samples."))
  }
}

```



# edgeR analysis

## CPM Filtering

```{r cpmfilter}
#| code-fold: true
#| code-summary: "🔍 CPM Filtering for edgeR"
#| message: true
#| warning: false

dge <- DGEList(counts = counts, samples = samples, group = samples$groupName)

cpm_mat <- cpm(dge, log = FALSE)
group <- dge$samples$group
keep_genes <- apply(cpm_mat, 1, function(row) {
  by_group <- split(row, group)
  group_pass <- sapply(by_group, function(vals) {
    mean(vals > params$edgeR_cpm_cutoff) >= params$edgeR_cpm_group_fraction
  })
  any(group_pass)
})

nendogeneous_genes_kept <- sum(keep_genes & is_endogenous)
# Keep ERCC genes if they are present .. no matter if they fail CPM test
if (isTRUE(params$useERCC)) {
  keep_genes <- keep_genes | is_nonzero_ercc
}

print(glue::glue("CPM filtering Criteria: "))
print(glue::glue("Atleast ", params$edgeR_cpm_group_fraction * 100, "% of samples in any group must have CPM > ", params$edgeR_cpm_cutoff))
print(glue::glue("🔍 Endogenous genes after filtering: ", nendogeneous_genes_kept))
# print(glue::glue("🔍 Filtered out ", sum(is_endogenous) - nendogeneous_genes_kept, " endogenous genes using CPM filtering. "))
dge_filtered <- dge[keep_genes, , keep.lib.sizes = FALSE]


# exit_gracefully()
```

## Apply Normalization

```{r setupdge}
#| code-fold: true
#| code-summary: "🧬 Set up DGE object"
#| message: true
#| warning: false

if (isTRUE(params$useERCC)) {
  message("✅ ERCC-based normalization enabled.")
  # Re-Identify ERCC rows in current DGE object .. these will all be non-zero already
  is_ercc <- grepl("^ERCC", rownames(dge_filtered))

  # Normalize using ERCC-only library sizes
  ercc_lib_sizes <- colSums(dge_filtered$counts[is_ercc, , drop = FALSE])
  dge_filtered$samples$norm.factors <- ercc_lib_sizes / exp(mean(log(ercc_lib_sizes)))

  # Create summary table
  ercc_norm_df <- data.frame(
    Sample = colnames(dge_filtered),
    ERCC_LibSize = format(ercc_lib_sizes, big.mark = ","),
    NormFactor = formatC(dge_filtered$samples$norm.factors, format = "f", digits = 3)
  )
  
  # Remove all ERCC rows before DE analysis
  dge_final <- dge_filtered[!is_ercc, , keep.lib.sizes = FALSE]
} else {
  message("ℹ️ Using TMM normalization.")
  dge_final <- calcNormFactors(dge_filtered)
}
```
```{r ercc-table, echo=FALSE}
#| code-fold: true
#| code-summary: "🧬 ERCC datatable"
#| message: true
#| warning: false
# Only run if normalization used ERCCs
if (isTRUE(params$useERCC)) {
  DT::datatable(
    ercc_norm_df,
    caption = "📊 ERCC-Based Library Sizes and Normalization Factors",
    options = list(pageLength = 20, autoWidth = TRUE)
  )
}
```

```{r normfactorplot}
#| code-fold: true
#| code-summary: "📊 Normalization Factor Plot"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false
dge_samples_df <- as.data.frame(dge_final$samples) %>% 
  rownames_to_column(var = "Sample") %>%
  mutate(
    Group = group,
    LibrarySize = lib.size,
    NormFactor = signif(norm.factors, 3)
  ) %>%
  dplyr::select(Sample, Group, LibrarySize, NormFactor)

# if original samples contains params$batch_column, add it to dge_samples_df
# Optionally add Batch to dge_samples_df
if (params$batch_column %in% colnames(samples)) {
  dge_samples_df$Batch <- selected_samples[[params$batch_column]][match(dge_samples_df$Sample, selected_samples[[params$sample_column]])]
  # Assign colors based on Batch
  dge_samples_df$Color <- group_colors[dge_samples_df$Group]


# Step 1: Define available shapes and assign per batch
available_shapes <- c("circle", "square", "diamond", "cross", "x", 
                      "triangle-up", "triangle-down", "triangle-left", 
                      "triangle-right", "star")

batches <- unique(dge_samples_df$Batch)
batch_shapes <- setNames(
  rep(available_shapes, length.out = length(batches)),
  batches
)

# Step 2: Create the plot
p <- plot_ly()

# Add real colored points (per group), no legend
p <- p %>% add_trace(
  data = dge_samples_df,
  x = ~LibrarySize,
  y = ~NormFactor,
  text = ~Sample,
  type = "scatter",
  mode = "markers",
  marker = list(
    color = dge_samples_df$Color,
    symbol = unname(batch_shapes[dge_samples_df$Batch])
  ),
  showlegend = FALSE
)

# Step 3: Add dummy black points for legend only
for (b in batches) {
  shape <- batch_shapes[b]
  p <- p %>% add_trace(
    x = 0, y = 0,  # Off-plot dummy point
    type = "scatter",
    mode = "markers",
    marker = list(
      color = "black",
      symbol = shape,
      size = 5
    ),
    name = b,
    showlegend = TRUE,
    hoverinfo = "none",
    legendgroup = b
  )
}

# Step 4: Layout
p <- p %>% layout(
  title = "Library Size vs Normalization Factor",
  xaxis = list(title = "Library Size"),
  yaxis = list(title = "Normalization Factor"),
  legend = list(title = list(text = "Batch"))
)

p
    
} else {

# Base plot setup
p <- plot_ly(dge_samples_df, 
             x = ~LibrarySize, 
             y = ~NormFactor, 
             text = ~Sample, 
             color = ~Group,
             colors = group_colors,
             type = "scatter", 
             mode = "markers")

# Final layout
p <- p %>% layout(
  title = "Library Size vs Normalization Factor",
  xaxis = list(title = "Library Size"),
  yaxis = list(title = "Normalization Factor")
)

p

}

# exit_gracefully()
```



## Sample Info

```{r sample_table}
#| code-fold: true
#| code-summary: "📋 Sample Metadata with Library Sizes and Norm Factors"
#| warning: false
#| message: false

# Extract sample info from dge object
sample_info <- dge_final$samples %>%
  tibble::rownames_to_column(var = "Sample") %>%
  mutate(
    Group = group,
    LibrarySize = format(lib.size, big.mark = ","),
    NormFactor = formatC(norm.factors, format = "f", digits = 3)
  )

if (params$batch_column %in% colnames(sample_info)) {
  sample_info <- sample_info %>%
    mutate(Batch = .[[params$batch_column]])
} else {
  sample_info <- sample_info %>%
    mutate(Batch = NA)
}

sample_info <- sample_info %>%
  dplyr::select(Sample, Group, LibrarySize, NormFactor, Batch)


DT::datatable(
  sample_info,
  options = list(
    pageLength = 20,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  ),
  rownames = FALSE
)

# exit_gracefully()

```

```{r save_normalized_counts}
#| code-fold: true
#| code-summary: "💾 Save Normalized Counts"
#| message: true
#| warning: false
# Save normalized counts to a file
normalized_counts <- cpm(dge_final, normalized.lib.sizes = TRUE, log = TRUE)
as.data.frame(normalized_counts) %>%
  rownames_to_column(var = "gene") -> normalized_counts
normalized_counts_file <- file.path(counts_dir, "edgeR_logCPM_counts.filtered.tsv")
write.table(normalized_counts, 
            file = normalized_counts_file, 
            sep = "\t", 
            quote = FALSE, 
            col.names = TRUE, 
            row.names = FALSE)
print(glue::glue("💾 Normalized counts saved to: ", normalized_counts_file))
# exit_gracefully()
```

```{r functions}
#| code-fold: true
#| code-summary: "🔧 Functions"
#| message: true
#| warning: false
#| echo: false

log2FC_threshold <- params$log2FC_threshold
pvalue_threshold <- params$pvalue_threshold
fdr_threshold <- params$fdr_threshold

make_violin_plot <- function(normalized_counts,
                             xlab = "Sample",
                             ylab = "Normalized Counts",
                             title = "Violin Plot of Normalized Counts by Sample") {
  # Prepare data
violin_data <- as.data.frame(normalized_counts) %>%
  # rownames_to_column(var = "gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "normalized_counts")

violin_data$group <- selected_samples[[params$group_column]][match(violin_data$sample, selected_samples[[params$sample_column]])]
violin_data$group <- factor(violin_data$group, levels = c(g1, g2))
violin_data$sample <- factor(violin_data$sample, levels = unique(violin_data$sample))
violin_data$normalized_counts <- as.numeric(violin_data$normalized_counts)
violin_data <- violin_data %>% filter(normalized_counts > 0)

# Plot
violin_plot <- ggplot(violin_data, aes(x = sample, y = normalized_counts, fill = group)) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = 16, outlier.size = 1, alpha = 0.5) +
  scale_y_log10() +
  scale_fill_manual(values = group_colors) +
  theme_bw() +
  labs(
    title = title,
    x = xlab,
    y = ylab,
    fill = "Group"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "top"
  ) + coord_flip()
return(violin_plot)
}

make_pca_plots <- function(normalized_counts){

normalized_counts %>% column_to_rownames(var = "gene") -> normalized_counts
# Perform PCA
# Remove constant genes (zero variance across samples)
normalized_counts <- normalized_counts[apply(normalized_counts, 1, function(x) var(x) > 0), ]

# Then run PCA
pca <- prcomp(t(normalized_counts), center = TRUE, scale. = TRUE)

pca_data <- as.data.frame(pca$x)

# Calculate % variance explained
pca_var <- pca$sdev^2
pca_var_exp <- round(100 * pca_var / sum(pca_var), 1)

# Add group and sample info
pca_data$sample <- rownames(pca_data)
pca_data$group <- selected_samples[[params$group_column]][match(pca_data$sample, selected_samples[[params$sample_column]])]
pca_data$group <- factor(pca_data$group, levels = c(g1, g2))

# Create custom hover text
pca_data$hover <- paste0(
  "Sample: ", pca_data$sample,
  "<br>Group: ", pca_data$group,
  "<br>PC1: ", round(pca_data$PC1, 2),
  "<br>PC2: ", round(pca_data$PC2, 2)
)

# Create interactive PCA plot
pca2d <- plot_ly(
  data = pca_data,
  x = ~PC1,
  y = ~PC2,
  color = ~group,
  colors = group_colors,
  text = ~hover,
  type = 'scatter',
  mode = 'markers',
  marker = list(size = 10),
  hoverinfo = "text"
) %>%
  layout(
    title = "PCA Plot of Normalized Counts",
    xaxis = list(title = paste0("PC1 (", pca_var_exp[1], "%)")),
    yaxis = list(title = paste0("PC2 (", pca_var_exp[2], "%)")),
    legend = list(orientation = "h", x = 0.3, y = -0.15)
  )

# Create hover text
pca_data$hover <- paste0(
  "Sample: ", pca_data$sample,
  "<br>Group: ", pca_data$group,
  "<br>PC1: ", round(pca_data$PC1, 2),
  "<br>PC2: ", round(pca_data$PC2, 2),
  "<br>PC3: ", round(pca_data$PC3, 2)
)

# 3D interactive PCA plot
pca3d <- plot_ly(
  data = pca_data,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~group,
  colors = group_colors,
  text = ~hover,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 5),
  hoverinfo = "text"
) %>%
  layout(
    title = "3D PCA Plot of Normalized Counts",
    scene = list(
      xaxis = list(title = paste0("PC1 (", pca_var_exp[1], "%)")),
      yaxis = list(title = paste0("PC2 (", pca_var_exp[2], "%)")),
      zaxis = list(title = paste0("PC3 (", pca_var_exp[3], "%)"))
    ),
    legend = list(orientation = "h", x = 0.3, y = -0.1)
  )
    return(list(
    pca2d = pca2d,
    pca3d = pca3d
  ))
}

make_volcano_plot <- function(volcano_df, 
                              pvalue_column_name = "PValue",
                              logFC_column_name = "logFC",
                              FDR_column_name = "FDR"
){

# remove all rows where Pvalue is NA
volcano_df <- volcano_df[!is.na(volcano_df[[pvalue_column_name]]), ]
# remove all rows where logFC is NA
volcano_df <- volcano_df[!is.na(volcano_df[[logFC_column_name]]), ]

# Rename logFC column to "logFC" if needed
if (logFC_column_name != "logFC") {
  names(volcano_df)[names(volcano_df) == logFC_column_name] <- "logFC"
}
if (pvalue_column_name != "PValue") {
  names(volcano_df)[names(volcano_df) == pvalue_column_name] <- "PValue"
}
if (FDR_column_name != "FDR") {
  names(volcano_df)[names(volcano_df) == FDR_column_name] <- "FDR"
}

volcano_df$logP <- -log10(volcano_df$PValue)

# Categorize each gene
volcano_df$Significance <- "NS"  # Default: not significant
volcano_df$Significance[volcano_df$PValue < pvalue_threshold & abs(volcano_df$logFC) >= log2FC_threshold] <- "p-value and log2FC"
volcano_df$Significance[volcano_df$PValue < pvalue_threshold & abs(volcano_df$logFC) < log2FC_threshold] <- "p-value"
volcano_df$Significance[volcano_df$PValue >= pvalue_threshold & abs(volcano_df$logFC) >= log2FC_threshold] <- "Log2FC"

# Define custom colors
sig_colors <- c(
  "NS" = "gray",
  "Log2FC" = "#009E73",             # Green
  "p-value" = "#0072B2",            # Blue
  "p-value and log2FC" = "red"  # Red
)

# Create hover text
volcano_df$hover <- paste0(
  "Gene: ", volcano_df$gene,
  "<br>logFC: ", round(volcano_df$logFC, 2),
  "<br>FDR: ", ifelse(is.na(volcano_df$FDR), NA, signif(volcano_df$FDR, 3)),
  "<br>PValue: ", signif(volcano_df$PValue, 3)
)

# Plot
volcano_plotly <- plot_ly(
  data = volcano_df,
  x = ~logFC,
  y = ~logP,
  type = "scatter",
  mode = "markers",
  color = ~Significance,
  colors = sig_colors,
  text = ~hover,
  hoverinfo = "text",
  marker = list(size = 6)
) %>%
  layout(
    title = list(text = paste0("Volcano Plot: ", g1, " vs ", g2)),
    xaxis = list(title = "log2 Fold Change"),
    yaxis = list(title = "-log10(P-value)"),
    legend = list(title = list(text = "Significance"), orientation = "h", x = 0.3, y = -0.2),
    shapes = list(
      # Horizontal line at -log10(0.05)
      list(type = "line", x0 = min(volcano_df$logFC), x1 = max(volcano_df$PValue),
           y0 = -log10(pvalue_threshold), y1 = -log10(pvalue_threshold),
           line = list(dash = "dot", width = 1, color = "black")),
      # Vertical line at logFC = -1
      list(type = "line", x0 = -1 * log2FC_threshold, x1 = -1 * log2FC_threshold,
           y0 = 0, y1 = max(volcano_df$logP),
           line = list(dash = "dot", width = 1, color = "black")),
      # Vertical line at logFC = 1
      list(type = "line", x0 = log2FC_threshold, x1 = log2FC_threshold,
           y0 = 0, y1 = max(volcano_df$logP),
           line = list(dash = "dot", width = 1, color = "black"))
    )
  )
return(volcano_plotly)
}

```


<!-- # 📊 Data Visualization -->
## Normalized counts Distribution plot

```{r plot_counts_violin_by_sample}
#| code-fold: true
#| code-summary: "🎻 Violin Plot of Normalized Counts by Sample (Colored by Group)"
#| fig.width: 10
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| warning: false

# Create violin plot
violin_plot <- make_violin_plot(normalized_counts,
                                 xlab = "Sample",
                                 ylab = "Normalized Counts (logCPM)",
                                 title = "Violin Plot of Normalized Counts by Sample (edgeR)")

print(violin_plot)
# exit_gracefully()
```


## 2-D PCA plot

```{r plot_pca_plotly}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

pcas <- make_pca_plots(normalized_counts)
pcas$pca2d

# exit_gracefully()
# print(pca_plot)
```


## 3-D PCA plot

```{r plot_pca_3d_plotly}
#| code-fold: true
#| code-summary: "🧬 3D PCA Plot"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false


pcas$pca3d
# exit_gracefully()
# print(pca_3d_plot)
```

## DEG results

```{r design}
#| code-fold: true
#| code-summary: "🧩 Design Matrix"


samples[[params$group_column]] <- relevel(factor(samples[[params$group_column]]), ref = "Uninf")
# Create design matrix
design <- model.matrix(as.formula(params$design_formula), data = samples)
# Report design matrix  dimensions
print(glue::glue("📊 Design matrix dimensions: ", nrow(design), " samples × ", ncol(design), " variables"))
# Check for singularities
if (any(is.na(design))) {
  stop("❌ Design matrix contains NA values. Please check your samplesheet and design formula.")
}

# Estimate dispersion
dge <- estimateDisp(dge_final, design)

# Fit GLM and run LRT
fit <- glmFit(dge, design)
lrt <- glmLRT(fit)

# Save DE table
top_table <- topTags(lrt, n = Inf)$table

# # Get test summary
# deg_results_stats_df <- summary(decideTests(lrt))
# as.data.frame(deg_results_stats_df) %>% rownames_to_column(var = "Group") -> deg_results_stats_df
# colnames(deg_results_stats_df) <- c("Group", "Direction", "dummy", "No. of genes")
# # Remove dummy column
# deg_results_stats_df <- deg_results_stats_df %>% dplyr::select(-dummy) %>% dplyr::select(-Group)

# Extract results table
res_table_edger <- topTags(lrt, n = Inf)$table

# Add average logCPM expression (from the DGE object)
res_table_edger$AveExpr <- rowMeans(cpm(dge, log = TRUE))

# Categorize gene regulation direction
res_table_edger$Direction <- dplyr::case_when(
  res_table_edger$FDR < fdr_threshold & res_table_edger$logFC > log2FC_threshold ~ "Up",
  res_table_edger$FDR < fdr_threshold & res_table_edger$logFC < -log2FC_threshold ~ "Down",
  TRUE ~ "NotSig"
)

# Create summary table
deg_results_stats_df <- as.data.frame(table(res_table_edger$Direction))
colnames(deg_results_stats_df) <- c("Direction", "No. of genes")

# Reorder rows to: Down, NotSig, Up
deg_results_stats_df <- deg_results_stats_df[match(c("Down", "NotSig", "Up"), deg_results_stats_df$Direction), ]

# write results table to file
results_file <- file.path(edger_deg_dir, "edgeR_results.tsv")
write.table(res_table_edger %>% rownames_to_column(var = "gene"), 
            file = results_file, 
            sep = "\t", 
            quote = FALSE, 
            col.names = TRUE, 
            row.names = FALSE)

print(glue::glue("💾 edgeR DE results saved to: ", results_file))
```

```{r deg_results_table}
#| code-fold: true
#| code-summary: "📊 DE Results Table"
#| message: true
#| warning: false

DT::datatable(
  deg_results_stats_df,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  ),
  rownames = FALSE
)
# exit_gracefully()
```

## MA Plot

```{r maplot}
#| code-fold: true
#| code-summary: "📈 MA Plot"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"


# Optionally, add a significance column for coloring
res_table_edger$Significant <- ifelse(res_table_edger$FDR < fdr_threshold, paste0("FDR < ",fdr_threshold), "Not Sig")

# Interactive MA plot
ma_plot <- plot_ly(
  data = res_table_edger,
  x = ~AveExpr,
  y = ~logFC,
  type = 'scatter',
  mode = 'markers',
  color = ~Significant,
  colors = c("red", "gray"),
  text = ~paste("Gene:", rownames(res_table_edger),
                "<br>logFC:", round(logFC, 2),
                "<br>FDR:", signif(FDR, 3)),
  hoverinfo = "text"
) %>%
  layout(
    title = "🧬 Interactive MA Plot",
    xaxis = list(title = "Average logCPM"),
    yaxis = list(title = "log2 Fold Change"),
    shapes = list(
      list(type = "line", x0 = min(res_table_edger$AveExpr), x1 = max(res_table_edger$AveExpr),
           y0 = 0, y1 = 0, line = list(dash = "dash", color = "black"))
    )
  )

# print plot
ma_plot
# exit_gracefully()
# print(ma_plot)
```

## Volcano Plot

```{r volcano_plotly_4color}
#| code-fold: true
#| code-summary: "🌋 Interactive Volcano Plot (4 colors like EnhancedVolcano)"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| warning: false


# Prepare data
volcano_df <- topTags(lrt, n = Inf)$table
volcano_df$gene <- rownames(volcano_df)
volcano_plotly <- make_volcano_plot(volcano_df,
                              pvalue_column_name = "PValue",
                              logFC_column_name = "logFC",
                              FDR_column_name = "FDR")


volcano_plotly
```

# DESeq2 analysis

## Low-count filtering

```{r lowcountfiltering}
#| code-fold: true
#| code-summary: "🧬 Low count filtering"
#| message: true
#| warning: false
#| echo: false

# # Rename columns only if they exist
# colnames(selected_samples)[colnames(selected_samples) == params$sample_column] <- "sample"
# colnames(selected_samples)[colnames(selected_samples) == params$group_column] <- "group"

# Apply DESeq2 low count filtering using params$DESeq2_low_count_cutoff and params$DESeq2_low_count_fraction
# Use endogenous counts for filtering
# A minimum of DESeq2_low_count_fraction fraction of samples from either group g1 or g2 should have counts > DESeq2_low_count_cutoff

# get original filtered counts
counts <- filtered_counts
is_ercc <- grepl("^ERCC", rownames(counts))
is_endogenous <- !is_ercc
is_nonzero_ercc <- is_ercc & (rowSums(counts) > 0)

group <- samples[[params$group_column]]

keep_genes_deseq2 <- apply(counts, 1, function(row) {
  by_group <- split(row, group)
  group_pass <- sapply(by_group, function(vals) {
    mean(vals > params$DESeq2_low_count_cutoff) >= params$DESeq2_low_count_group_fraction
  })
  any(group_pass)
})

if (isTRUE(params$useERCC)) {
  keep_genes_deseq2 <- keep_genes_deseq2 | is_nonzero_ercc
}

nendogeneous_genes_kept <- sum(keep_genes_deseq2 & is_endogenous)
nercc_genes_kept <- sum(keep_genes_deseq2 & is_nonzero_ercc)
print(glue::glue("Low-count filtering Criteria: "))
print(glue::glue("Atleast ", params$DESeq2_low_count_group_fraction * 100, "% of samples in any group must have count > ", params$DESeq2_low_count_cutoff))
print(glue::glue("✅ Endogenous genes after filtering: ", nendogeneous_genes_kept))
# if (isTRUE(params$useERCC)) {
# print(glue::glue("✅ ERCC genes after filtering low count genes for DESeq2 analysis: ", nercc_genes_kept))
# }
```

```{r deseq2_normalization}
#| code-fold: true
#| code-summary: "🧬 DESEq2 Normalization"
#| message: true
#| warning: false
#| echo: false

design_formula <- as.formula(params$design_formula)
dds <- DESeqDataSetFromMatrix(countData = counts[keep_genes_deseq2, ], 
                              colData = samples, 
                              design = design_formula)

if (isTRUE(params$useERCC)) {
  print(glue("🧪 ERCC scaling will be applied"))
  
  ercc_rows <- grepl("^ERCC", rownames(dds))
  # Estimate size factors using ERCC-only rows
  dds <- estimateSizeFactors(dds, controlGenes = ercc_rows)

  # Remove ERCC rows from dds after normalization
  dds <- dds[!ercc_rows, ]
} else {
  print(glue("📏 Default DESeq2 size factor normalization"))
  dds <- estimateSizeFactors(dds)
}

# Pull normalized library size (sizeFactors) and original library size
dds_samples_df <- as.data.frame(colData(dds)) %>%
  rownames_to_column(var = "Sample") %>%
  mutate(
    Group = groupName,
    LibrarySize = colSums(counts(dds)),
    NormFactor = signif(sizeFactors(dds), 3)
  ) %>%
  dplyr::select(Sample, Group, LibrarySize, NormFactor)

# Add Batch column and colors
if (params$batch && params$batch_column %in% colnames(samples)) {
  dds_samples_df$Batch <- selected_samples[[params$batch_column]][match(dds_samples_df$Sample, selected_samples[[params$sample_column]])]
  dds_samples_df$Color <- group_colors[dds_samples_df$Group]

  # Define shapes per batch
  available_shapes <- c("circle", "square", "diamond", "cross", "x", 
                        "triangle-up", "triangle-down", "triangle-left", 
                        "triangle-right", "star")
  batches <- unique(dds_samples_df$Batch)
  batch_shapes <- setNames(
    rep(available_shapes, length.out = length(batches)),
    batches
  )

  # Base plot
  p <- plot_ly()

  # Real points (color by group, shape by batch)
  p <- p %>% add_trace(
    data = dds_samples_df,
    x = ~LibrarySize,
    y = ~NormFactor,
    text = ~Sample,
    type = "scatter",
    mode = "markers",
    marker = list(
      color = dds_samples_df$Color,
      symbol = unname(batch_shapes[dds_samples_df$Batch]),
      size = 6
    ),
    name="", # prevents 'trace 0' from showing in hover
    showlegend = FALSE
  )

  # Dummy points for legend (black, shape only)
  for (b in batches) {
    shape <- batch_shapes[b]
    p <- p %>% add_trace(
      x = 0, y = 0,
      type = "scatter",
      mode = "markers",
      marker = list(
        color = "black",
        symbol = shape,
        size = 6
      ),
      name = b,
      showlegend = TRUE,
      hoverinfo = "none",
      legendgroup = b
    )
  }

  # Layout
  p <- p %>% layout(
    title = "Library Size vs Normalization Factor (DESeq2)",
    xaxis = list(title = "Library Size"),
    yaxis = list(title = "Size Factor"),
    legend = list(title = list(text = "Batch"))
  )
} else {
  # Simple plot without batch encoding
  p <- plot_ly(dds_samples_df,
               x = ~LibrarySize,
               y = ~NormFactor,
               text = ~Sample,
               color = ~Group,
               colors = group_colors,
               type = "scatter",
               mode = "markers",
               marker = list(size = 6))

  p <- p %>% layout(
    title = "Library Size vs Normalization Factor (DESeq2)",
    xaxis = list(title = "Library Size"),
    yaxis = list(title = "Size Factor")
  )
}

p

```

```{r deseq2_sampleinfo}
#| code-fold: true
#| code-summary: "📋 DESeq2 Sample Metadata with Library Sizes and Norm Factors"
#| warning: false
#| message: false
#| echo: false

# Extract sample info from DESeq2 object
sample_info_deseq2 <- as.data.frame(colData(dds)) %>%
  rownames_to_column(var = "Sample") %>%
  mutate(
    Group = group,
    LibrarySize = format(colSums(counts(dds)), big.mark = ","),
    NormFactor = formatC(sizeFactors(dds), format = "f", digits = 3),
    Batch = if ("Batch" %in% names(.)) Batch 
          else if ("batch" %in% names(.)) batch 
          else NA
  ) %>%
  dplyr::select(Sample, Group, LibrarySize, NormFactor, Batch)

DT::datatable(
  sample_info_deseq2,
  options = list(
    pageLength = 20,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  ),
  rownames = FALSE
)
# exit_gracefully()
```

```{r save_deseq2_normalized_counts}
#| code-fold: true
#| code-summary: "💾 Save DESeq2 Normalized Counts"
#| message: true
#| warning: false
#| echo: false

# Save DESeq2 normalized counts
normalized_counts_deseq2 <- counts(dds, normalized = TRUE)
as.data.frame(normalized_counts_deseq2) %>%
  rownames_to_column(var = "gene") -> normalized_counts_deseq2
normalized_counts_deseq2_file <- file.path(counts_dir, "DESeq2_normalized_counts.tsv")
write.table(normalized_counts_deseq2, 
            file = normalized_counts_deseq2_file, 
            sep = "\t", 
            quote = FALSE, 
            col.names = TRUE, 
            row.names = FALSE)
print(glue::glue("💾 DESeq2 normalized counts saved to: ", normalized_counts_deseq2_file))
```

## Normalized counts Distribution plot

```{r plot_counts_violin_by_sample_deseq2}
#| code-fold: true
#| code-summary: "🎻 Violin Plot of Normalized Counts by Sample (Colored by Group)"
#| fig.width: 10
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"

# Create violin plot
violin_plot_deseq2 <- make_violin_plot(normalized_counts_deseq2,
                                        xlab = "Sample",
                                        ylab = "Normalized Counts",
                                        title = "Violin Plot of Normalized Counts by Sample (DESeq2)")
                                        

print(violin_plot_deseq2)
# exit_gracefully()                            
```


## 2-D PCA plot

```{r plot_pca_plotly_deseq2}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

pcas_deseq2 <- make_pca_plots(normalized_counts_deseq2)
pcas_deseq2$pca2d

# exit_gracefully()
# print(pca_plot)
```


## 3-D PCA plot

```{r plot_pca_3d_plotly_deseq2}
#| code-fold: true
#| code-summary: "🧬 3D PCA Plot"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false


pcas_deseq2$pca3d
# exit_gracefully()
# print(pca_3d_plot)
```

## DEG results

```{r deseq2_results}
#| code-fold: true
#| code-summary: "🧩 DESeq2 Results"
#| message: true
#| warning: false
#| echo: false

# Run DESeq2 analysis
dds <- DESeq(dds)

# Extract the variable name from formula string
design_variable <- trimws(gsub("~", "", params$design_formula))

# Construct contrast vector
contrast_vector <- c(design_variable, g1, g2)

# Run DESeq2 results with contrast
res <- results(dds, contrast = contrast_vector)
# Add gene names
res$gene <- rownames(res)
# Save DESeq2 results to a file
res_file <- file.path(deseq2_deg_dir, "DESeq2_results.tsv")
write.table(res, 
            file = res_file, 
            sep = "\t", 
            quote = FALSE, 
            col.names = TRUE, 
            row.names = TRUE)
print(glue::glue("💾 DESeq2 results saved to: ", res_file))

# Assume 'res' is your DESeq2 results object
res_table_deseq2 <- as.data.frame(res)

# Add gene IDs as a column (if rownames are gene IDs)
res_table_deseq2$gene <- rownames(res_table_deseq2)

# Categorize gene direction
res_table_deseq2$Direction <- case_when(
  !is.na(res_table_deseq2$padj) & res_table_deseq2$padj < fdr_threshold & res_table_deseq2$log2FoldChange >  log2FC_threshold ~ "Up",
  !is.na(res_table_deseq2$padj) & res_table_deseq2$padj < fdr_threshold & res_table_deseq2$log2FoldChange < -log2FC_threshold ~ "Down",
  TRUE ~ "NotSig"
)

# Create summary stats table
deg_results_stats_df_deseq2 <- as.data.frame(table(res_table_deseq2$Direction))
colnames(deg_results_stats_df_deseq2) <- c("Direction", "No. of genes")

# Reorder
deg_results_stats_df_deseq2 <- deg_results_stats_df_deseq2[match(c("Down", "NotSig", "Up"), deg_results_stats_df_deseq2$Direction), ]
```

```{r show_deseq2_results}
#| code-fold: true
#| code-summary: "📊 DESeq2 Results Summary"
#| message: false
#| warning: false
#| echo: false

# Show interactive table
DT::datatable(deg_results_stats_df_deseq2, rownames = FALSE, 
              options = list(
                pageLength = 10,
                autoWidth = TRUE,
                columnDefs = list(list(className = 'dt-center', targets = "_all"))))

```

```{r volcano_plotly_deseq2}
#| code-fold: true
#| code-summary: "🌋 Interactive Volcano Plot (DESeq2)"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"

# Prepare data
volcano_df_deseq2 <- res_table_deseq2
volcano_df_deseq2$gene <- rownames(volcano_df_deseq2)
volcano_plotly <- make_volcano_plot(volcano_df_deseq2,
                              pvalue_column_name = "pvalue",
                              logFC_column_name = "log2FoldChange",
                              FDR_column_name = "padj")
#
volcano_plotly
```


# Combined DEG Results

```{r combined_results}
#| code-fold: true
#| code-summary: "🧬 Combined DEG Results"
#| message: true
#| warning: false
#| echo: false

if (!"gene" %in% colnames(res_table_edger)) {
  res_table_edger <- res_table_edger %>% rownames_to_column(var = "gene")
}
names(res_table_edger)[names(res_table_edger) == "logFC"] <- "log2FoldChange"


if (!"gene" %in% colnames(res_table_deseq2)) {
  res_table_deseq2 <- res_table_deseq2 %>% rownames_to_column(var = "gene")
}
names(res_table_deseq2)[names(res_table_deseq2) == "pvalue"] <- "PValue"
names(res_table_deseq2)[names(res_table_deseq2) == "padj"] <- "FDR"

res_table_edger <- res_table_edger %>%
  dplyr::select(gene, log2FoldChange, PValue, FDR, AveExpr, Direction)

res_table_deseq2 <- res_table_deseq2 %>%
  dplyr::select(gene, log2FoldChange, PValue, FDR, baseMean, Direction)

# Combine results from edgeR and DESeq2
combined_results <- full_join(res_table_edger, res_table_deseq2, by = "gene", suffix = c("_edgeR", "_DESeq2"))
combined_results[combined_results == ""] <- NA

names(combined_results)[names(combined_results) == "AveExpr"] <- "AveExpr_edgeR"
names(combined_results)[names(combined_results) == "baseMean"] <- "baseMean_DESeq2"

output_file <- file.path(output_base, "combined_deg_results.tsv")
write.table(combined_results, 
            file = output_file, 
            sep = "\t", 
            quote = FALSE, 
            col.names = TRUE, 
            row.names = FALSE)
print(glue::glue("💾 Combined results saved to: ", output_file))
# Show interactive table

combined_results$Direction_edgeR <- as.character(combined_results$Direction_edgeR)
combined_results$Direction_DESeq2 <- as.character(combined_results$Direction_DESeq2)

# Assuming combined_results is your data frame
deg_comparison_table <- combined_results %>%
  dplyr::count(Direction_edgeR, Direction_DESeq2, name = "Count") %>%
  arrange(desc(Count))

deg_comparison_table$Match <- deg_comparison_table$Direction_edgeR == deg_comparison_table$Direction_DESeq2

# Display with highlighting where directions match
DT::datatable(deg_comparison_table, 
          rownames = FALSE,
          options = list(pageLength = 10,
                         autoWidth = TRUE, 
                         ordering = FALSE, 
                         columnDefs = list(list(className = 'dt-center',visible = FALSE, targets = 3)))) %>%
  formatStyle(
    columns = "Match",
    target = "row",
    backgroundColor = styleEqual(c(TRUE, FALSE), c("#D4EDDA", NA))
  )

# Only consider rows where Match is TRUE or FALSE (exclude NA)
valid_rows <- !is.na(deg_comparison_table$Match)

total <- sum(deg_comparison_table$Count[valid_rows])
matched <- sum(deg_comparison_table$Count[deg_comparison_table$Match == TRUE],na.rm=TRUE)

concordance_percent <- round(100 * matched / total, 2)

cat("✅ edgeR-DESeq2 Concordance: ", concordance_percent, "%\n")
```

```{r Venn}
#| code-fold: true
#| code-summary: "🧬 Venn Diagram of DEGs"
#| fig.width: 8
#| fig.height: 8
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| warning: false

# Create sets
edgeR_up <- combined_results %>% filter(Direction_edgeR == "Up") %>% pull(gene)
edgeR_down <- combined_results %>% filter(Direction_edgeR == "Down") %>% pull(gene)
deseq2_up <- combined_results %>% filter(Direction_DESeq2 == "Up") %>% pull(gene)
deseq2_down <- combined_results %>% filter(Direction_DESeq2 == "Down") %>% pull(gene)

# Generate Venn diagram
venn.plot <- draw.quad.venn(
  area1 = length(edgeR_up),
  area2 = length(edgeR_down),
  area3 = length(deseq2_up),
  area4 = length(deseq2_down),
  n12 = length(intersect(edgeR_up, edgeR_down)),
  n13 = length(intersect(edgeR_up, deseq2_up)),
  n14 = length(intersect(edgeR_up, deseq2_down)),
  n23 = length(intersect(edgeR_down, deseq2_up)),
  n24 = length(intersect(edgeR_down, deseq2_down)),
  n34 = length(intersect(deseq2_up, deseq2_down)),
  n123 = length(Reduce(intersect, list(edgeR_up, edgeR_down, deseq2_up))),
  n124 = length(Reduce(intersect, list(edgeR_up, edgeR_down, deseq2_down))),
  n134 = length(Reduce(intersect, list(edgeR_up, deseq2_up, deseq2_down))),
  n234 = length(Reduce(intersect, list(edgeR_down, deseq2_up, deseq2_down))),
  n1234 = length(Reduce(intersect, list(edgeR_up, edgeR_down, deseq2_up, deseq2_down))),
  category = c("edgeR Up", "edgeR Down", "DESeq2 Up", "DESeq2 Down"),
  fill = c("tomato", "skyblue", "lightgreen", "orange"),
  cex = 1.2,
  cat.cex = 1.2
)
grid.draw(venn.plot)
```

<!-- ```{r exit} -->
<!-- exit_gracefully() -->
<!-- print(glue::glue(" Exiting early: Debugging script intentionally halted.")) -->
<!-- ``` -->

