---
title: "Create normalized counts"
format:
  html:
    theme: flatly
    toc: true
    toc-location: left
    toc-depth: 4
execute:
  echo: false
  warning: false
  message: true
params:
  counts_file: "/Users/vishal/Documents/GitRepos/DiffEx/.test/counts_matrix.tsv"
  samplesheet: "/Users/vishal/Documents/GitRepos/DiffEx/.test/samples.tsv"
  # test dataset 2
  # counts_file: "/Users/vishal/Documents/Data/Analysis/20250909_RNAseq-2/DEG/counts_matrix.tsv"
  # samplesheet: "/Users/vishal/Documents/Data/Analysis/20250909_RNAseq-2/DEG/samples.tsv"
  useERCC: FALSE
  ercc_mix: 1 # choices: 1, 2
  # group1: "Lyt_DBR1"
  # group2: "Lyt_NTC"
  sample_column: "sampleName"
  group_column: "groupName"
  usebatch: FALSE
  batch_column: "batch"
  outdir: "/Users/vishal/Desktop/20250909_RNAseq-2/DEG/DiffEx_results/host/usebatch/Lyt_DBR1_vs_Lyt_NTC"
  host: "Hs" # choices: "Hs", "Mm"
  genes_selection: "host" # choices: "host", "virus", "both" # filter out genes other than this selection
  # test dataset 1
  # counts_file: "~/Documents/Data/Analysis/20250423_RNAseq-1/counts/counts_matrix.tsv"
  # samplesheet: "~/Documents/Data/Analysis/20250423_RNAseq-1/counts/samples.tsv"
  # useERCC: FALSE
  # ercc_mix: 1 # choices: 1, 2
  # group1: "GFP48h"
  # group2: "UT"
  # sample_column: "sampleName"
  # group_column: "groupName" # DEG will always be based of this column
  # usebatch: FALSE # use batch as covariate in design matrix
  # batch_column: "batch"
  # outdir: "~/Documents/Data/Analysis/20250423_RNAseq-1/DiffEx_results"
  # host: "Mm" # choices: "Hs", "Mm"
  # genes_selection: "host" # choices: "host", "virus", "both" # filter out genes other than this selection
  # thresholds and cutoffs
  # log2FC_threshold: 1.0 # use 1.0 and not 1 .. .to make it explicitly float for wrapping python package
  # pvalue_threshold: 0.05
  # fdr_threshold: 0.05
  edgeR_cpm_cutoff: 0.1 # use float to make it explicitly float for wrapping python package
  edgeR_cpm_group_fraction: 0.5
  DESeq2_low_count_cutoff: 2
  DESeq2_low_count_group_fraction: 0.5

---

# 🔬 Differential Expression Analysis

This report performs standard DGE analysis using input count data, metadata, and user-defined contrasts.

---
```{r library}
#| code-fold: true
#| code-summary: "📦 Load Required Packages"
suppressPackageStartupMessages({
  library(limma)
  library(edgeR)
  library(DESeq2)
  library(tibble)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(plotly)
  library(EnhancedVolcano)
  library(knitr)
  library(kableExtra)
  library(ggplotify)
  library(ggpubr)
  library(ggrepel)
  library(openxlsx)
  library(glue)
  library(DT)
  library(ComplexHeatmap)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
  library(msigdbr)
  library(UpSetR)
  library(ggplot2)
  library(plotly)
  library(openxlsx)
  library(yaml)
  library(fs)
  library(VennDiagram)
  library(grid)
  library(sva)
  library(UpSetR)
})

exit_gracefully <- function() {
  message("Exiting early: Debugging script intentionally halted.")
  knitr::knit_exit()
}
# Paths
output_base <- params$outdir
# gsea_dir <- file.path(output_base, "gsea")

# Create folders
fs::dir_create(c(output_base))

# write out params
# Convert Quarto parameters (params) to a named list
yaml_params <- as.list(params)

# Write to YAML file
yaml::write_yaml(yaml_params, file.path(output_base, "params.yaml"))
```

## Reading inputs

```{r readinput}
#| code-fold: true
#| code-summary: "📥 Read Count Matrix and Sample Sheet"
#| message: true
#| warning: false

print(glue::glue("📥 Reading input files: "))
print(glue::glue(params$counts_file, " and "))
print(glue::glue(params$samplesheet))

# Read the sample sheet
samplesheet <- params$samplesheet
samples <- read.table(samplesheet, sep = "\t", header = TRUE, check.names = FALSE)
keep_cols <- c(params$sample_column, params$group_column)
if (isTRUE(params$usebatch)) {
  keep_cols <- c(keep_cols, params$batch_column)
  if (!params$batch_column %in% colnames(samples)) {
    stop(glue::glue("❌ Batch column '{params$batch_column}' not found in samplesheet."))
  }
}
# Find which of the required columns are actually present
present_cols <- intersect(keep_cols, colnames(samples))

# Check: all required columns must be present
if (length(present_cols) != length(keep_cols)) {
  missing_cols <- setdiff(keep_cols, present_cols)
  stop("❌ The following required columns are missing from 'samples': ", paste(missing_cols, collapse = ", "))
}


# Read the counts matrix
counts_data <- read.table(params$counts_file,
                     check.names = FALSE,  # prevent automatic conversion of hyphen to dot
                     row.names = 1, 
                     sep = "\t", 
                     header = TRUE)
colnames(counts_data) <- gsub("-", "_", colnames(counts_data))
counts_metadata <- counts_data[, c( "species", grep("^gene_", colnames(counts_data), value = TRUE))]
sample_columns <- samples[[params$sample_column]]
counts <- counts_data[, c(sample_columns)]
original_counts <- counts

# Report original dimensions
print(glue::glue("📊 Counts matrix loaded: ", nrow(counts), " genes (rows) × ", ncol(counts), " samples (columns)"))

counts_dir <- file.path(output_base)
```



## Filtering counts matrix by genes selection

```{r genes_selection}
#| code-fold: true
#| code-summary: "🔍 Filter by genes_selection"
#| message: true
#| warning: false

print(glue::glue("Genes selection: ", params$genes_selection))
print(glue::glue("useERCC: ", params$useERCC))
print(glue::glue("usebatch: ", params$usebatch))

# Filter out genes based on the genes_selection
if (params$genes_selection == "host") {
  if (isTRUE(params$useERCC)) {
    # Select only host and ERCC genes
    if (params$host == "Hs") {
      filtered_counts <- counts[grepl("^ENSG|^ERCC", rownames(counts)), ]
    } else if (params$host == "Mm") {
      filtered_counts <- counts[grepl("^ENSMUSG|^ERCC", rownames(counts)), ]
    }
  } else {
    # Select only host genes no ERCC genes
    if (params$host == "Hs") {
      filtered_counts <- counts[grepl("^ENSG", rownames(counts)), ]
    } else if (params$host == "Mm") {
      filtered_counts <- counts[grepl("^ENSMUSG", rownames(counts)), ]
    }
  }
  nercc_genes <- nrow(filtered_counts[grepl("^ERCC", rownames(filtered_counts)), ])
  print(glue::glue("🦠 Host-only genes selected: ", nrow(filtered_counts), " genes remaining (including ", nercc_genes, " ERCC genes)"))
} else if (params$genes_selection == "virus") {
  if (isTRUE(params$useERCC)) {
    # Filter out host genes .. keep ERCC
    filtered_counts <- counts[!grepl("^ENS", rownames(counts)), ]
  } else {
    # Filter out host genes and ERCC genes
    filtered_counts <- counts[!grepl("^ERCC|^ENS", rownames(counts)), ]
  }
  nercc_genes <- nrow(filtered_counts[grepl("^ERCC", rownames(filtered_counts)), ])
  print(glue::glue("🦠 Virus-only genes selected: ", nrow(filtered_counts), " genes remaining (including ", nercc_genes, " ERCC genes)"))
} else if (params$genes_selection == "both") {
  if (!isTRUE(params$useERCC)) {
    # Keep both host and virus genes .. remove ERCC
    filtered_counts <- counts[grepl("^ERCC", rownames(counts)), ]
  }
  nercc_genes <- nrow(filtered_counts[grepl("^ERCC", rownames(filtered_counts)), ])
  print(glue::glue("🦠 Both host and virus genes included: ", nrow(filtered_counts), " genes remaining (including ", nercc_genes, " ERCC genes)"))
} else {
  stop("❌ Invalid mode selected. Choose 'host', 'virus', or 'both'.")
}
# Report dimensions
print(glue::glue("📦 Filtered counts matrix: ", nrow(filtered_counts), " genes × ", ncol(filtered_counts), " samples"))

counts <- filtered_counts
ready_for_deg_raw_counts <- counts[rowSums(counts) != 0,]
n_all_zero_genes <- nrow(counts) - nrow(ready_for_deg_raw_counts)
if (n_all_zero_genes > 0) {
  print(glue::glue("❌ Warning: ", n_all_zero_genes, " genes have zero counts in all samples and will be removed."))
} else {
  print("✅ No all-zero genes found.")
}
library_sizes <- colSums(ready_for_deg_raw_counts)

# remove ERCCs genes which have zero reads in any sample
is_ercc <- grepl("^ERCC", rownames(ready_for_deg_raw_counts))
ercc_counts <- ready_for_deg_raw_counts[is_ercc, ]
nonzero_ercc_rows <- apply(ercc_counts, 1, function(x) all(x != 0))
is_nonzero_ercc <- rownames(ready_for_deg_raw_counts) %in% rownames(ercc_counts)[nonzero_ercc_rows]
remove_ercc <- is_ercc & !is_nonzero_ercc

# update with selected nonzero ERCC genes
ready_for_deg_raw_counts <- ready_for_deg_raw_counts[!remove_ercc, ]
is_ercc <- grepl("^ERCC", rownames(ready_for_deg_raw_counts))
is_endogenous <- !is_ercc


if (isTRUE(params$useERCC)) {
  print(glue::glue("❌ Warning: ",sum(remove_ercc), " ERCC genes have ZERO counts in 1 or more samples and will be excluded!."))

  # Check if any ERCC genes have zero counts in all samples
  if (sum(is_nonzero_ercc) == 0) {
    stop("❌ All ERCC genes have zero counts in all samples. Please check your data or set useERCC to FALSE.")
  }
  print(glue::glue("✅ ", sum(is_nonzero_ercc), " ERCC genes have non-zero counts in all samples."))
} else {
  # useERCC is set to FALSE
  # Check if any ERCC genes have zero counts in all samples
  if (sum(is_ercc) > 0) {
    print(glue::glue("❌ ",sum(is_ercc)," ERCC genes are present but will be removed before analysis."))
  }
  ready_for_deg_raw_counts <- ready_for_deg_raw_counts[!is_ercc, ]
}
# Report dimensions
print(glue::glue("📦 Newly filtered counts matrix: ", nrow(ready_for_deg_raw_counts), " genes × ", ncol(ready_for_deg_raw_counts), " samples"))
```


```{r ercc_calculations}
#| code-fold: true
#| code-summary: "🧬 ERCC Calculations"
#| message: true
#| warning: false
#| echo: false

if (isTRUE(params$useERCC)) {

  # ercc_conc <- read.csv("/Users/vishal/Documents/GitRepos/DiffEx/diffex/assets/ERCC92_conc.txt", row.names = 1)
  # TODO: make this txt file part of DiffEx package
  # ercc_url <- "https://raw.githubusercontent.com/mschertzer/ercc_analysis/refs/heads/master/ERCC92_conc.txt"
  ercc_url <- "https://raw.githubusercontent.com/dremellab/DiffEx/refs/heads/main/diffex/assets/ERCC92_conc.txt"
  
  # Read the tab-delimited ERCC file directly from GitHub
  ercc_conc <- read.delim(ercc_url, check.names = FALSE)
  colnames(ercc_conc) <- c("sortID",
                           "ERCC_ID",
                           "subgroup",
                           "Mix1_conc",
                           "Mix2_conc",
                           "FC_ratio",
                           "log2FC")
  
  # Filter the ERCC concentration data for the selected mix
  if (params$ercc_mix == 1) {
    ercc_conc <- ercc_conc[c("ERCC_ID", "Mix1_conc")]
  } else if (params$ercc_mix == 2) {
    ercc_conc <- ercc_conc[c("ERCC_ID", "Mix2_conc")]
  }
  colnames(ercc_conc) <- c("ERCC_ID", "ercc_conc")
  # make ERCC_ID the rownames after converting hyphen to underscore
  ercc_conc$ERCC_ID <- gsub("-", "_", ercc_conc$ERCC_ID)
  as.data.frame(ercc_conc) %>% column_to_rownames(var = "ERCC_ID") -> ercc_conc
  # calculated molecules from concentration
  ercc_conc$log2molecules <- log(ercc_conc$ercc_conc / 1000 * 20 * 6.022e5,2)
  
  counts_df <- as.data.frame(ready_for_deg_raw_counts)
  counts_df$gene_id <- rownames(counts_df)
  
  counts_data$gene_id <- rownames(counts_data)
  
  merged_counts <- merge(counts_df,counts_data[,c("gene_id", "gene_length_kb")], by = "gene_id")
  as.data.frame(merged_counts) %>% column_to_rownames(var = "gene_id") -> merged_counts
  # merged_counts now has gene_length_kb as a column
  
    # Identify expression-only columns
  expr_cols <- setdiff(colnames(merged_counts), "gene_length_kb")
  
  # Compute RPKM
  rpkm_counts <- sweep(merged_counts[, expr_cols], 1, merged_counts$gene_length_kb, "/")  # divide by gene length
  rpkm_counts <- sweep(rpkm_counts, 2, library_sizes / 1e6, "/")  # divide by library size (in millions)

  # Log2-transform RPKM
  log2_rpkm_counts <- log2(rpkm_counts)

  # Subset ERCC rows
  ercc_rows <- grepl("^ERCC", rownames(log2_rpkm_counts))
  log2_ercc_counts <- log2_rpkm_counts[ercc_rows, ]
  
  # Remove -Inf and NA
  log2_ercc_counts <- log2_ercc_counts[!apply(log2_ercc_counts, 1, function(x) any(is.na(x) | is.infinite(x))), ]
  
  # Clean up rownames: keep only the first part before the '|'
  rownames(log2_ercc_counts) <- sub("\\|.*", "", rownames(log2_ercc_counts))

  common_erccs <- intersect(rownames(log2_ercc_counts), rownames(ercc_conc))
  
  counts_matched <- log2_ercc_counts[common_erccs, ]
  # reorder ercc_conc to match counts_matched
  ercc_conc <- ercc_conc[common_erccs, ]

  # Initialize results data frame
  ercc_qc <- data.frame(
    Sample = colnames(counts_matched),
    Slope = NA_real_,
    Intercept = NA_real_,
    Rsquared = NA_real_
  )
  
  # Fit linear model per sample
  for (s in colnames(counts_matched)) {
    y <- counts_matched[,s] # already log2 transformed RPKM values
    x <- ercc_conc$log2molecules
  
    fit <- lm(y ~ x)
    ercc_qc[ercc_qc$Sample == s, "Slope"]     <- coef(fit)[2]
    ercc_qc[ercc_qc$Sample == s, "Intercept"] <- coef(fit)[1]
    ercc_qc[ercc_qc$Sample == s, "Rsquared"]  <- summary(fit)$r.squared
  }

  # Create summary table
  ercc_norm_df <- data.frame(
    Sample = ercc_qc$Sample,
    Slope = format(ercc_qc$Slope, digits = 3),
    Intercept = format(ercc_qc$Intercept, digits = 3),
    Rsquared = format(ercc_qc$Rsquared, digits = 3)
  )
  
# Implement corrected log2RPKM = (log2RPKM - intercept) / slope (per sample)
pseudo_log2_normalized_counts <- log2_rpkm_counts

for (s in colnames(pseudo_log2_normalized_counts)) {
  intercept <- ercc_qc$Intercept[ercc_qc$Sample == s]
  slope <- ercc_qc$Slope[ercc_qc$Sample == s]

  # Sanity check
  if (length(intercept) != 1 || length(slope) != 1) {
    stop(glue::glue("❌ Could not uniquely match slope/intercept for sample {s}."))
  }

  # Apply transformation
  pseudo_log2_normalized_counts[, s] <- (pseudo_log2_normalized_counts[, s] - intercept) / slope
}

  
  pseudo_log2_normalized_counts[pseudo_log2_normalized_counts == -Inf] <- NA
}

# pseudo_log2_normalized_counts_df <- as.data.frame(pseudo_log2_normalized_counts) %>%
#   rownames_to_column(var = "gene_id")
# 
# write.table(pseudo_log2_normalized_counts_df,
#             file = file.path(output_base, "ERCC_pseudo_log2normalized_counts.tsv"),
#             sep = "\t",
#             quote = FALSE,
#             col.names = TRUE,
#             row.names = FALSE)

```

```{r ercc_table}
#| code-fold: true
#| code-summary: "📊 ERCC Summary Table"
#| message: true
#| warning: false
#| echo: false

if (isTRUE(params$useERCC)) {
  DT::datatable(
    ercc_norm_df,
    caption = "📊 ERCC-Based Library Sizes and Intercept/Slope/R-squared",
    rownames = FALSE,
    options = list(pageLength = 20, autoWidth = TRUE)
  )
}
```


```{r functions}
#| code-fold: true
#| code-summary: "🔧 Functions"
#| message: true
#| warning: false
#| echo: false

# Adds a gene_id column from rownames if not already present:
add_gene_id_column_if_missing <- function(df) {
  if (!"gene_id" %in% colnames(df)) {
    df$gene_id <- rownames(df)
  }
  return(df)
}

# Moves the gene_id column to rownames if it exists:
set_gene_id_as_rownames_if_exists <- function(df) {
  if ("gene_id" %in% colnames(df)) {
    rownames(df) <- df$gene_id
    df$gene_id <- NULL
  }
  return(df)
}


log2FC_threshold <- params$log2FC_threshold
pvalue_threshold <- params$pvalue_threshold
fdr_threshold <- params$fdr_threshold

make_violin_plot <- function(normalized_counts,
                             xlab = "Sample",
                             ylab = "Normalized Counts",
                             title = "Violin Plot of Normalized Counts by Sample") {

# Prepare data
violin_data <- as.data.frame(normalized_counts)

# Remove 'gene' column if it exists
if ("gene" %in% colnames(violin_data)) {
  violin_data$gene <- NULL
}

# Proceed to reshape data
violin_data <- violin_data %>%
  pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "normalized_counts"
  )

# violin_data$group <- selected_samples[[params$group_column]][match(violin_data$sample, selected_samples[[params$sample_column]])]
violin_data$group <- factor(violin_data$group)
violin_data$group <- factor(samples$groupName)
violin_data$sample <- factor(violin_data$sample, 
  levels = violin_data %>%
    distinct(sample, group) %>%
    arrange(group, sample) %>%
    pull(sample))
violin_data$normalized_counts <- as.numeric(violin_data$normalized_counts)
violin_data <- violin_data %>% filter(normalized_counts > 0)

# Plot
violin_plot <- ggplot(violin_data, aes(x = sample, y = normalized_counts, fill = group)) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = 16, outlier.size = 1, alpha = 0.5) +
  scale_y_log10() +
  scale_fill_manual(values = group_colors) +
  theme_bw() +
  labs(
    title = title,
    x = xlab,
    y = ylab,
    fill = "Group"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "top"
  ) + coord_flip()
return(violin_plot)
}

make_pca_plots <- function(normalized_counts,
                           selected_samples,
                           title = "PCA Plot of Normalized Counts") {

  # Convert input to data frame and set rownames
  normalized_counts <- normalized_counts %>%
    as.data.frame() %>%
    {
      if ("gene" %in% colnames(.)) {
        column_to_rownames(., var = "gene")
      } else {
        .
      }
    }

  # Remove genes with zero variance
  normalized_counts <- normalized_counts[apply(normalized_counts, 1, var) > 0, ]

  # Perform PCA
  pca <- prcomp(t(normalized_counts), center = TRUE, scale. = TRUE)
  pca_data <- as.data.frame(pca$x)

  # % Variance explained
  pca_var <- pca$sdev^2
  pca_var_exp <- round(100 * pca_var / sum(pca_var), 1)

  # Annotate PCA metadata
  pca_data$sample <- rownames(pca_data)
  pca_data$group <- selected_samples[[params$group_column]][match(pca_data$sample, selected_samples[[params$sample_column]])]
  pca_data$group <- factor(pca_data$group)

  if (isTRUE(params$usebatch)) {
    pca_data$batch <- selected_samples[[params$batch_column]][match(pca_data$sample, selected_samples[[params$sample_column]])]
    pca_data$batch <- factor(pca_data$batch)
  } else {
    pca_data$batch <- NA
  }

  # Hover text for 2D
  pca_data$hover2d <- paste0(
    "Sample: ", pca_data$sample,
    "<br>Group: ", pca_data$group,
    "<br>Batch: ", pca_data$batch,
    "<br>PC1: ", round(pca_data$PC1, 2),
    "<br>PC2: ", round(pca_data$PC2, 2)
  )

  # Hover text for 3D
  pca_data$hover3d <- paste0(
    "Sample: ", pca_data$sample,
    "<br>Group: ", pca_data$group,
    "<br>Batch: ", pca_data$batch,
    "<br>PC1: ", round(pca_data$PC1, 2),
    "<br>PC2: ", round(pca_data$PC2, 2),
    "<br>PC3: ", round(pca_data$PC3, 2)
  )

  # 2D PCA plot
  pca2d <- plot_ly(
    data = pca_data,
    x = ~PC1,
    y = ~PC2,
    color = ~group,
    text = ~hover2d,
    type = 'scatter',
    mode = 'markers',
    symbol = if (isTRUE(params$usebatch)) ~batch else I("circle"),
    marker = list(size = 10),
    hoverinfo = "text"
  ) %>%
    layout(
      title = title,
      xaxis = list(title = paste0("PC1 (", pca_var_exp[1], "%)")),
      yaxis = list(title = paste0("PC2 (", pca_var_exp[2], "%)")),
      legend = list(
        orientation = "v",   # vertical legend
        x = 1.05,            # move to right outside plot
        y = 1,
        font = list(size = 10),
        itemsizing = 'constant'
      ),
      margin = list(r = 150) # add room on right for legend
    )


  # 3D PCA plot
  pca3d <- plot_ly(
  data = pca_data,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~group,
  text = ~hover3d,
  type = "scatter3d",
  mode = "markers",
  symbol = if (isTRUE(params$usebatch)) ~batch else I("circle"),
  marker = list(size = 5),
  hoverinfo = "text"
) %>%
  layout(
    title = title,
    scene = list(
      xaxis = list(title = paste0("PC1 (", pca_var_exp[1], "%)")),
      yaxis = list(title = paste0("PC2 (", pca_var_exp[2], "%)")),
      zaxis = list(title = paste0("PC3 (", pca_var_exp[3], "%)"))
    ),
    legend = list(
      orientation = "v",   # vertical layout
      x = 1.05,            # push legend to the right
      y = 1,
      font = list(size = 10),
      itemsizing = 'constant'
    ),
    margin = list(r = 150)  # increase right margin
  )

  return(list(
    pca2d = pca2d,
    pca3d = pca3d
  ))
}


plot_all_violin_by_group <- function(expr_df, samples_df, xlab="", ylab="",title = "Gene expression distribution") {
  library(ggplot2)
  library(reshape2)
  
  # Melt expression matrix into long format
  df_long <- melt(expr_df, id.vars = "gene",
                  variable.name = "sampleName", value.name = "expression")
  
  # Merge with sample group info
  df_long <- merge(df_long, samples_df, by = "sampleName")
  
  # Create horizontal violin plot
  p <- ggplot(df_long, aes(x = sampleName, y = expression, fill = groupName)) +
    geom_violin(scale = "width", trim = FALSE) +
    coord_flip() +
    theme_bw(base_size = 14) +
    theme(
      axis.text.y = element_text(size = 10),
      axis.text.x = element_text(size = 12),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_blank(),
      legend.title = element_blank(),
      plot.title = element_text(hjust = 0.5, size = 16)
    ) +
    labs(
      title = title,
      x = xlab,
      y = ylab
    )
  
  return(p)
}

```



# limma analysis

## Normalization

```{r limma}
#| code-fold: true
#| code-summary: "📊 Limma Analysis"
#| message: true
#| warning: false
#| echo: false

  group_design <- model.matrix(~ 0 + samples[[params$group_column]])
  colnames(group_design) <- levels(samples[[params$group_column]])
  if (isTRUE(params$usebatch)) {
  # Convert to factors
  samples$batch <- factor(samples$batch)
  samples$groupName <- factor(samples$groupName)
  
  # Create a design matrix including both batch and group
  limma_design <- model.matrix(~ 0 + batch + groupName, data = samples)
  
  # Optional: cleaner column names
  colnames(limma_design) <- make.names(colnames(limma_design))
  
  # Sanity check
  stopifnot(ncol(filtered_counts) == nrow(limma_design))
  } else {
    # directly use group_design
    limma_design <- group_design
  }


if (isTRUE(params$useERCC)) {
  
  print(glue::glue("✅ ERCC standard curve slope - based normalization enabled."))
  dims <- dim(pseudo_log2_normalized_counts)
  
  # get batch corrected counts
  if (isTRUE(params$usebatch)){
    pseudo_log2_normalized_counts_batch_corrected <- limma::removeBatchEffect(
      pseudo_log2_normalized_counts,
      batch = samples[[params$batch_column]],
      design = model.matrix(~ 0 + samples[[params$group_column]])
    )
    pseudo_log2_normalized_counts_batch_corrected <- as.data.frame(pseudo_log2_normalized_counts_batch_corrected)
  }
  
} else {
  print(glue::glue("✅ Using Limma-voom for normalization."))

  # 1. Create DGEList
  dge <- DGEList(counts = ready_for_deg_raw_counts)
  group <- samples[[params$group_column]]
  dge$samples$group <- group
  if (isTRUE(params$usebatch)) {
    batch <- samples[[params$batch_column]]
    dge$samples$batch <- batch
  }
  
  # 2. Filtering
  keep <- filterByExpr(dge, group = group)
  dge <- dge[keep,, keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge)  

  # 3. Design matrix with optional batch covariate
  # using limma_design
  
  # 4. Voom
  v <- voom(dge, limma_design, plot = TRUE)
  
  # 5. Save log2 values
  log2_raw_cpm <- as.data.frame(cpm(dge, log = TRUE, prior.count = 1))
  log2_normalized_counts_limma <- as.data.frame(v$E)
  dims <- dim(log2_normalized_counts_limma)
  
  if (isTRUE(params$usebatch)){
    log2_normalized_counts_limma_batch_corrected <- limma::removeBatchEffect(
      log2_normalized_counts_limma,
      batch = samples[[params$batch_column]],
      design = model.matrix(~ 0 + samples[[params$group_column]])
    )
    log2_normalized_counts_limma_batch_corrected <- as.data.frame(log2_normalized_counts_limma_batch_corrected)
  }

}

```

## Sample Info

```{r limma_normfactor_table, echo=FALSE}
#| code-fold: true
#| code-summary: "🧬 datatable"
#| message: true
#| warning: false
# Only run if normalization used ERCCs

if (isTRUE(params$useERCC)) {
  print(glue::glue("📊 log2 normalized pseudo RPKM counts matrix dimensions: {dims[1]} genes × {dims[2]} samples"))
} else {
  print(glue::glue("📊 log2 normalized counts matrix dimensions post filterByExpr: {dims[1]} genes × {dims[2]} samples"))
  limma_sampleinfo_df <- as.data.frame(dge$samples) %>%
    rownames_to_column(var = "Sample") %>%
    mutate(
      Group = group,
      LibrarySize = lib.size,
      NormFactor = signif(norm.factors, 3),
      Batch = if ("batch" %in% colnames(.)) batch else NA
    ) %>%
    dplyr::select(Sample, Group, any_of("Batch"),LibrarySize, NormFactor) %>%
    dplyr::select(where(~ !all(is.na(.))))
  
  DT::datatable(
    limma_sampleinfo_df,
    caption = "📊 Library Sizes and Normalization Factors",
    rownames = FALSE,
    options = list(pageLength = 20, autoWidth = TRUE)
  )
}
```

```{r save_limma_normalized_counts}
#| code-fold: true
#| code-summary: "💾 Save Normalized Counts"
#| message: true
#| warning: false


if (isTRUE(params$useERCC)) {
  normalized_counts_file <- file.path(counts_dir, "limma_log2normalized_pseudo_rpkm_counts.tsv")
  
    # Save normalized counts to a file
  as.data.frame(pseudo_log2_normalized_counts) %>%
    rownames_to_column(var = "gene") -> pseudo_log2_normalized_counts_for_output
  write.table(pseudo_log2_normalized_counts_for_output,
              file = normalized_counts_file,
              sep = "\t",
              quote = FALSE,
              col.names = TRUE,
              row.names = FALSE)
  print(glue::glue("💾 Normalized counts saved to: ", normalized_counts_file))
  
  pseudo_rpkm_counts <- 2^pseudo_log2_normalized_counts/20000
  pseudo_rpkm_counts[is.na(pseudo_rpkm_counts)] <- 0
  pseudo_rpkm_counts_for_output <- as.data.frame(pseudo_rpkm_counts) %>%
    rownames_to_column(var = "gene")
  pseudo_rpkm_counts_file <- file.path(counts_dir, "limma_pseudo_rpkm_counts.tsv")
  write.table(pseudo_rpkm_counts_for_output,
              file = pseudo_rpkm_counts_file,
              sep = "\t",
              quote = FALSE,
              col.names = TRUE,
              row.names = FALSE)
  if (isTRUE(params$usebatch)) {
    pseudo_log2_normalized_counts_batch_corrected_file <- file.path(counts_dir, "limma_log2normalized_pseudo_rpkm_counts_batch_corrected.tsv")
    as.data.frame(pseudo_log2_normalized_counts_batch_corrected) %>%
      rownames_to_column(var = "gene") -> pseudo_log2_normalized_counts_batch_corrected_for_output
    write.table(pseudo_log2_normalized_counts_batch_corrected_for_output,
                file = pseudo_log2_normalized_counts_batch_corrected_file,
                sep = "\t",
                quote = FALSE,
                col.names = TRUE,
                row.names = FALSE)
    print(glue::glue("💾 Batch corrected normalized counts saved to: ", pseudo_log2_normalized_counts_batch_corrected_file))
  }
} else {
  normalized_counts_file <- file.path(counts_dir, "limma_voom_normalized_counts.tsv")
  
  as.data.frame(log2_normalized_counts_limma) %>%
    rownames_to_column(var = "gene") -> log2_normalized_counts_limma_for_output
  write.table(log2_normalized_counts_limma_for_output,
              file = normalized_counts_file,
              sep = "\t",
              quote = FALSE,
              col.names = TRUE,
              row.names = FALSE)
  print(glue::glue("💾 Normalized counts saved to: ", normalized_counts_file))
  
  if (isTRUE(params$usebatch)){
    log2_normalized_counts_limma_batch_corrected_file <- file.path(counts_dir, "limma_voom_normalized_counts_batch_corrected.tsv")
    as.data.frame(log2_normalized_counts_limma_batch_corrected) %>%
      rownames_to_column(var = "gene") -> log2_normalized_counts_limma_batch_corrected_for_output
    write.table(log2_normalized_counts_limma_batch_corrected_for_output,
                file = log2_normalized_counts_limma_batch_corrected_file,
                sep = "\t",
                quote = FALSE,
                col.names = TRUE,
                row.names = FALSE)
    print(glue::glue("💾 Batch corrected normalized counts saved to: ", log2_normalized_counts_limma_batch_corrected_file))
  }

}
# exit_gracefully()
```

## Counts Distribution

::: {.panel-tabset}

### Before normalization
```{r limma_violin_before_norm}
#| code-fold: true
#| code-summary: "📊 Violin Plot of Normalized Counts"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"

if (isTRUE(params$useERCC)) {
  log2_rpkm_counts_for_plots <- log2_rpkm_counts[!apply(log2_rpkm_counts, 1, function(x) any(is.na(x) | is.infinite(x))), ]
as.data.frame(log2_rpkm_counts_for_plots) %>% rownames_to_column(var = "gene") -> log2_rpkm_counts_for_plots
  violin_plot <- plot_all_violin_by_group(log2_rpkm_counts_for_plots,
                                          as.data.frame(samples),
                                   xlab = "Sample",
                                   ylab = "Log2 RPKM Raw Counts")
} else {
  log2_raw_cpm %>% rownames_to_column(var = "gene") -> log2_raw_cpm_for_plots
  violin_plot <- plot_all_violin_by_group(log2_raw_cpm_for_plots,
                                          as.data.frame(samples),
                                   xlab = "Sample",
                                   ylab = "Log2 Raw CPM Counts",
                                   title = "")
}
violin_plot
```
### After normalization
```{r limma_violins}
#| code-fold: true
#| code-summary: "📊 Violin Plot of Normalized Counts"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"


# Create violin plot of normalized counts
if (isTRUE(params$useERCC)) {
  if (isTRUE(params$usebatch)) {
    pseudo_log2_normalized_counts_batch_corrected_for_plots <- pseudo_log2_normalized_counts_batch_corrected[!apply(pseudo_log2_normalized_counts_batch_corrected, 1, function(x) any(is.na(x) | is.infinite(x))), ]
    as.data.frame(pseudo_log2_normalized_counts_batch_corrected_for_plots) %>% rownames_to_column(var = "gene") -> pseudo_log2_normalized_counts_batch_corrected_for_plots
    ylabel <- "Log2 Normalized Pseudo RPKM Counts (Batch Corrected)"
    violin_plot <- plot_all_violin_by_group(pseudo_log2_normalized_counts_batch_corrected_for_plots,
                                            as.data.frame(samples),
                                     xlab = "Sample",
                                     ylab = ylabel,
                                     title = "")
  } else {
    pseudo_log2_normalized_counts_for_plots <- pseudo_log2_normalized_counts[!apply(pseudo_log2_normalized_counts, 1, function(x) any(is.na(x) | is.infinite(x))), ]
    as.data.frame(pseudo_log2_normalized_counts_for_plots) %>% rownames_to_column(var = "gene") -> pseudo_log2_normalized_counts_for_plots
    ylabel <- "Log2 Normalized Pseudo RPKM Counts"
    violin_plot <- plot_all_violin_by_group(pseudo_log2_normalized_counts_for_plots,
                                    as.data.frame(samples),
                                     xlab = "Sample",
                                     ylab = ylabel,
                                     title = "")
  }
} else {
  if (isTRUE(params$usebatch)) {
    log2_normalized_counts_limma_batch_corrected %>% rownames_to_column(var = "gene") -> log2_normalized_counts_limma_batch_corrected_for_plots
    ylabel <- "Log2 Normalized Counts (Batch Corrected)"
    violin_plot <- plot_all_violin_by_group(log2_normalized_counts_limma_batch_corrected_for_plots,
                                    as.data.frame(samples),
                               xlab = "Sample",
                               ylab = ylabel,
                               title = "")

  } else {
    log2_normalized_counts_limma %>% rownames_to_column(var = "gene") -> log2_normalized_counts_limma_for_plots
    ylabel <- "Log2 Normalized Counts"
    violin_plot <- plot_all_violin_by_group(log2_normalized_counts_limma_for_plots,
                                    as.data.frame(samples),
                                 xlab = "Sample",
                                 ylab = ylabel,
                                 title = "")
  }
}
violin_plot
```
:::

## 2-D PCA plot

::: {.panel-tabset}

### Before normalization

```{r plot_pca_before_limma}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

if (isTRUE(params$useERCC)) {
  pcas_before <- make_pca_plots(log2_rpkm_counts_for_plots,
                                as.data.frame(samples),
                               title = "PCA Plot of Log2 RPKM Counts")
} else {
pcas_before <- make_pca_plots(log2_raw_cpm_for_plots,
                              as.data.frame(samples),
                               title = "PCA Plot of Log2 Raw CPM Counts")
}

pcas_before$pca2d

# exit_gracefully()
# print(pca_plot)
```

### After normalization

```{r plot_pca_after_limma}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false
if (isTRUE(params$useERCC)) {
    if (isTRUE(params$usebatch)) {
    pcas_after <- make_pca_plots(pseudo_log2_normalized_counts_batch_corrected_for_plots,
                                 as.data.frame(samples),
                               title = "PCA Plot of Log2 Normalized Pseudo RPKM Counts (Batch corrected)")
  } else {
    pcas_after <- make_pca_plots(pseudo_log2_normalized_counts_for_plots,as.data.frame(samples),
                               title = "PCA Plot of Log2 Normalized Pseudo RPKM Counts")
  }  
} else {
    if (isTRUE(params$usebatch)) {
    pcas_after <- make_pca_plots(log2_normalized_counts_limma_batch_corrected_for_plots,as.data.frame(samples),
                               title = "PCA Plot of voom Normalized Counts (Batch corrected)")
  } else {
    pcas_after <- make_pca_plots(log2_normalized_counts_limma_for_plots,as.data.frame(samples),
                               title = "PCA Plot of voom Normalized Counts")
  }
}
pcas_after$pca2d
```

:::

## 3-D PCA plot

::: {.panel-tabset}

### Before normalization

```{r plot_3dpca_before_limma}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

pcas_before$pca3d

# exit_gracefully()
# print(pca_plot)
```

### After normalization

```{r plot_3dpca_after_limma}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

pcas_after$pca3d

# exit_gracefully()
# print(pca_plot)
```

:::


# edgeR analysis

## CPM Filtering

```{r cpmfilter}
#| code-fold: true
#| code-summary: "🔍 CPM Filtering for edgeR"
#| message: true
#| warning: false

if (isTRUE(params$useERCC)) {
  ready_for_deg_raw_counts <- round(pseudo_rpkm_counts)
}

dge_edger <- DGEList(counts = ready_for_deg_raw_counts, samples = samples, group = samples[[params$group_column]])

cpm_mat <- cpm(dge_edger, log = FALSE)
group <- dge_edger$samples$group
keep_genes <- apply(cpm_mat, 1, function(row) {
  by_group <- split(row, group)
  group_pass <- sapply(by_group, function(vals) {
    mean(vals > params$edgeR_cpm_cutoff) >= params$edgeR_cpm_group_fraction
  })
  any(group_pass)
})

dge_filtered <- dge_edger[keep_genes, , keep.lib.sizes = FALSE]

ercc_genes_kept <- grepl("^ERCC",rownames(dge_filtered))
nercc_genes_kept <- sum(ercc_genes_kept)
nendogenous_genes_kept <- sum(keep_genes) - nercc_genes_kept

print(glue::glue("CPM filtering Criteria: "))
print(glue::glue("Atleast ", params$edgeR_cpm_group_fraction * 100, "% of samples in any group must have CPM > ", params$edgeR_cpm_cutoff))
print(glue::glue("🔍 Endogenous genes after filtering: ", nendogenous_genes_kept))
print(glue::glue("🔍 ERCC genes after filtering: ", nercc_genes_kept))

# exit_gracefully()
```

## Apply Normalization

```{r setup_edger}
#| code-fold: true
#| code-summary: "🧬 Set up DGE object"
#| message: true
#| warning: false

dge_final <- dge_filtered
dge_final <- calcNormFactors(dge_filtered)

if (isTRUE(params$useERCC)) {
  print(glue::glue("✅ Using ERCC-normalized pseudo RPKM counts from limma analysis."))
  dge_final$samples$norm.factors <- rep(1, ncol(dge_final))
} else {
  print(glue::glue("ℹ️ Using TMM normalization."))
}
```

## Sample Info

```{r edger_normfactor_table, echo=FALSE}
#| code-fold: true
#| code-summary: "🧬 ERCC datatable"
#| message: true
#| warning: false
# Only run if normalization used ERCCs

  edger_sampleinfo_df <- as.data.frame(dge_final$samples) %>%
      rownames_to_column(var = "Sample") %>%
      mutate(
        Group = group,
        LibrarySize = lib.size,
        NormFactor = signif(norm.factors, 3),
        Batch = if ("batch" %in% colnames(.)) batch else NA
  ) %>%
    dplyr::select(Sample, Group, any_of("Batch"),LibrarySize, NormFactor) %>%
    dplyr::select(where(~ !all(is.na(.))))
  
  DT::datatable(
    edger_sampleinfo_df,
    caption = "📊 Library Sizes and Normalization Factors",
    rownames = FALSE,
    options = list(pageLength = 20, autoWidth = TRUE)
  )
```

```{r edger_save_normalized_counts}
#| code-fold: true
#| code-summary: "💾 Save Normalized Counts"
#| message: true
#| warning: false
# Save normalized counts to a file
normalized_counts_edger <- as.data.frame(cpm(dge_final, normalized.lib.sizes = TRUE, log = TRUE))
normalized_counts_edger %>%
  rownames_to_column(var = "gene") -> normalized_counts_edger_for_plots
normalized_counts_file <- file.path(counts_dir, "edgeR_TMM_normalized_logCPM_counts.tsv")
write.table(normalized_counts_edger_for_plots,
            file = normalized_counts_file,
            sep = "\t",
            quote = FALSE,
            col.names = TRUE,
            row.names = FALSE)
print(glue::glue("💾 Normalized counts saved to: ", normalized_counts_file))

if (isTRUE(params$usebatch)) {

  # Get batch and optional biological condition
  batch <- samples[[params$batch_column]]
  group <- samples[[params$group_column]]
  mod <- model.matrix(~ group)  # Preserve biological signal
  
  batch_counts <- table(batch)
  batch_group_counts <- table(batch, group)
  
  if (any(batch_counts < 2) || any(batch_group_counts < 2)) {
    print(glue::glue("⚠️ Warning: Some batches or batch-group combinations have less than 2 samples. ComBat cannot be run."))
  } else {
  
  # Apply ComBat on normalized logCPM
  normalized_counts_edger_batch_corrected <- as.data.frame(sva::ComBat(dat = normalized_counts_edger, batch = batch, mod = mod, par.prior = TRUE))
  normalized_counts_edger_batch_corrected %>%
    rownames_to_column(var = "gene") -> normalized_counts_edger_for_plots
  normalized_counts_file_batch_corrected <- file.path(counts_dir, "edgeR_TMM_normalized_logCPM_counts_batch_corrected.tsv")
  write.table(normalized_counts_edger_for_plots,
              file = normalized_counts_file_batch_corrected,
              sep = "\t",
              quote = FALSE,
              col.names = TRUE,
              row.names = FALSE)
  print(glue::glue("💾 Batch corrected normalized counts saved to: ", normalized_counts_file_batch_corrected))
  }
}

# exit_gracefully()
```

## Counts Distribution

::: {.panel-tabset}

### Before normalization

```{r violin_before}
#| code-fold: true
#| code-summary: "🎻 Violin Plot of Raw Counts by Sample (Colored by Group)"
#| fig.width: 10
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| warning: false


# Before normalization (assumes raw counts, norm.factors = 1)
dge_raw <- dge_final
dge_raw$samples$norm.factors <- 1
logcpm_before <- cpm(dge_raw, log = TRUE)
logcpm_before <- logcpm_before %>% as.data.frame() %>% tibble::rownames_to_column("gene")

# Create violin plot
violin_plot <- plot_all_violin_by_group(logcpm_before,
                                 as.data.frame(samples),
                                 xlab = "Sample",
                                 ylab = "Counts (logCPM)",
                                 title = "")

print(violin_plot)
# exit_gracefully()
```

### After Normalization

```{r violin_after}
#| code-fold: true
#| code-summary: "🎻 Violin Plot of Raw Counts by Sample (Colored by Group)"
#| fig.width: 10
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| warning: false

if (isTRUE(params$usebatch)) {
  ylabel="Normalized Counts (logCPM batch corrected)"
} else {
  ylabel="Normalized Counts (logCPM)"
}
# Create violin plot
violin_plot <- plot_all_violin_by_group(normalized_counts_edger_for_plots,
                                        as.data.frame(samples),
                               xlab = "Sample",
                               ylab = ylabel,
                               title = "")
print(violin_plot)
```

:::

## 2-D PCA plot

::: {.panel-tabset}

### Before normalization

```{r edger_plot_pca_before}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

pcas_before <- make_pca_plots(logcpm_before,
                             as.data.frame(samples),
                               title = "PCA Plot of Log2 Raw CPM Counts")
pcas_before$pca2d

# exit_gracefully()
# print(pca_plot)
```

### After normalization

```{r edger_plot_pca_after}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

if (isTRUE(params$usebatch)) {
  pcas_after <- make_pca_plots(normalized_counts_edger_for_plots,
                               as.data.frame(samples),
                               title = "PCA Plot of Log2 Normalized Counts (Batch corrected)")
} else {
  pcas_after <- make_pca_plots(normalized_counts_edger_for_plots,
                               as.data.frame(samples),
                               title = "PCA Plot of Log2 Normalized Counts")
}
pcas_after$pca2d

# exit_gracefully()
# print(pca_plot)
```

:::

## 3-D PCA plot

:::{.panel-tabset}

### Before normalization

```{r plot_pca_3d_before}
#| code-fold: true
#| code-summary: "🧬 3D PCA Plot"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false


pcas_before$pca3d
# exit_gracefully()
# print(pca_3d_plot)
```

### After normalization

```{r plot_pca_3d_after}
#| code-fold: true
#| code-summary: "🧬 3D PCA Plot"
#| fig.width: 8
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false


pcas_after$pca3d
# exit_gracefully()
# print(pca_3d_plot)
```

:::


# DESeq2 analysis

## Low-count filtering

```{r deseq2_lowcountfiltering}
#| code-fold: true
#| code-summary: "🧬 Low count filtering"
#| message: true
#| warning: false
#| echo: false


# get original filtered counts
if (isTRUE(params$useERCC)) {
  ready_for_deg_raw_counts <- round(pseudo_rpkm_counts)
} 

is_ercc <- grepl("^ERCC", rownames(ready_for_deg_raw_counts))
is_endogenous <- !is_ercc
is_nonzero_ercc <- is_ercc & (rowSums(ready_for_deg_raw_counts) > 0)

group <- samples[[params$group_column]]

keep_genes_deseq2 <- apply(ready_for_deg_raw_counts, 1, function(row) {
  by_group <- split(row, group)
  group_pass <- sapply(by_group, function(vals) {
    mean(vals > params$DESeq2_low_count_cutoff) >= params$DESeq2_low_count_group_fraction
  })
  any(group_pass)
})

if (isTRUE(params$useERCC)) {
  keep_genes_deseq2 <- keep_genes_deseq2 | is_nonzero_ercc
}

nendogeneous_genes_kept <- sum(keep_genes_deseq2 & is_endogenous)
nercc_genes_kept <- sum(keep_genes_deseq2 & is_nonzero_ercc)
print(glue::glue("Low-count filtering Criteria: "))
print(glue::glue("Atleast ", params$DESeq2_low_count_group_fraction * 100, "% of samples in any group must have count > ", params$DESeq2_low_count_cutoff))
print(glue::glue("✅ Endogenous genes after filtering: ", nendogeneous_genes_kept))

```

## Normalization

```{r deseq2_normalization}
#| code-fold: true
#| code-summary: "🧬 DESEq2 Normalization"
#| message: true
#| warning: false
#| echo: false

# build egder design
if (isTRUE(params$usebatch)){
  model_str <- paste("~",params$batch_column,"+",params$group_column)
} else {
  model_str <- paste("~", params$group_column)
}
design_formula <- as.formula(model_str)
dds <- DESeqDataSetFromMatrix(countData = ready_for_deg_raw_counts[keep_genes_deseq2, ], 
                              colData = samples, 
                              design = design_formula)

if (isTRUE(params$useERCC)) {
  print(glue("🧪 ERCC slope based normalization will be applied"))
  sizeFactors(dds) <- rep(1, ncol(dds))
} else {
  print(glue("📏 Default DESeq2 size factor normalization"))
  dds <- estimateSizeFactors(dds)
}


```

## Sample Info

```{r deseq2_sampleinfo}
#| code-fold: true
#| code-summary: "📋 DESeq2 Sample Metadata with Library Sizes and Norm Factors"
#| warning: false
#| message: false
#| echo: false

  # Pull normalized library size (sizeFactors) and original library size
  deseq2_sampleinfo_df <- as.data.frame(colData(dds)) %>%
  rownames_to_column(var = "Sample") %>%
  mutate(
    Group = groupName,
    LibrarySize = colSums(counts(dds)),
    NormFactor = signif(sizeFactors(dds), 3),
    Batch = if ("batch" %in% colnames(.)) batch else NA
  ) %>%
    dplyr::select(Sample, Group, any_of("Batch"),LibrarySize, NormFactor) %>%
    dplyr::select(where(~ !all(is.na(.))))
  
  DT::datatable(
  deseq2_sampleinfo_df,
  options = list(
    pageLength = 20,
    autoWidth = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  ),
  rownames = FALSE
)
# exit_gracefully()
```

```{r save_deseq2_normalized_counts}
#| code-fold: true
#| code-summary: "💾 Save DESeq2 Normalized Counts"
#| message: true
#| warning: false
#| echo: false


# Run VST and extract assay
vst_obj <- vst(dds, blind = TRUE)
normalized_counts_deseq2 <- assay(vst_obj) %>% as.data.frame()

# Add gene column only if not present
normalized_counts_deseq2_for_plots <- normalized_counts_deseq2 %>%
  tibble::rownames_to_column(var = "gene")

# Define output file path
normalized_counts_deseq2_file <- file.path(counts_dir, "DESeq2_vst_normalized_counts.tsv")

# Write to file
write.table(normalized_counts_deseq2_for_plots,
            file = normalized_counts_deseq2_file,
            sep = "\t",
            quote = FALSE,
            col.names = TRUE,
            row.names = FALSE)

print(glue::glue("💾 DESeq2 normalized counts saved to: {normalized_counts_deseq2_file}"))

if (isTRUE(params$usebatch)){
  
  # Get batch and optional biological condition
  batch <- samples[[params$batch_column]]
  group <- samples[[params$group_column]]
  mod <- model.matrix(~ group)  # Preserve biological signal

  batch_counts <- table(batch)
  batch_group_counts <- table(batch, group)
  
  if (any(batch_counts < 2) || any(batch_group_counts < 2)) {
    print(glue::glue("⚠️ Warning: Some batches or batch-group combinations have less than 2 samples. ComBat cannot be run."))
  } else {
  
  # Apply ComBat on normalized VST counts
  normalized_counts_deseq2_batch_corrected <- as.data.frame(sva::ComBat(dat = normalized_counts_deseq2, batch = batch, mod = mod, par.prior = TRUE))
  normalized_counts_deseq2_for_plots <- normalized_counts_deseq2_batch_corrected %>%
    tibble::rownames_to_column(var = "gene")
  
  # Define output file path
  normalized_counts_deseq2_file_batch_corrected <- file.path(counts_dir, "DESeq2_vst_normalized_counts_batch_corrected.tsv")
  
  # Write to file
  write.table(normalized_counts_deseq2_for_plots,
              file = normalized_counts_deseq2_file_batch_corrected,
              sep = "\t",
              quote = FALSE,
              col.names = TRUE,
              row.names = FALSE)
  
  print(glue::glue("💾 DESeq2 batch corrected normalized counts saved to: {normalized_counts_deseq2_file_batch_corrected}"))
  }
}

```

## Counts Distribution

:::{.panel-tabset}

### Before normalization

```{r violin_deseq2_before}
#| code-fold: true
#| code-summary: "🎻 Violin Plot of Raw Counts by Sample (Colored by Group)"
#| fig.width: 10
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

# BEFORE normalization (raw counts)
raw_counts <- counts(dds, normalized = FALSE)
logcounts_before <- log2(raw_counts + 1)
logcounts_before_df <- as.data.frame(logcounts_before) %>% tibble::rownames_to_column("gene")


# Create violin plot
violin_plot_deseq2 <- plot_all_violin_by_group(logcounts_before_df,
                                               as.data.frame(samples),
                                        xlab = "Sample",
                                        ylab = "Raw Counts(log2)",
                                        title = "")
                                        

print(violin_plot_deseq2)
# exit_gracefully()                            
```

### After normalization

```{r violin_deseq2_after}
#| code-fold: true
#| code-summary: "🎻 Violin Plot of VST Counts by Sample (Colored by Group)"
#| fig.width: 10
#| fig.height: 10
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

if (isTRUE(params$usebatch)) {
  ylabel="VST Normalized Counts (Batch corrected)"
} else {
  ylabel="VST Normalized Counts"
}

# Create violin plot
violin_plot_deseq2 <- plot_all_violin_by_group(normalized_counts_deseq2_for_plots,
                                       as.data.frame(samples),
                                        xlab = "Sample",
                                        ylab = ylabel,
                                        title = "")
                                        

print(violin_plot_deseq2)
# exit_gracefully()                            
```

:::

## 2-D PCA plot

::: {.panel-tabset}

### Before normalization

```{r plot_pca_plotly_deseq2_before}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

pcas_deseq2_before <- make_pca_plots(logcounts_before_df,as.data.frame(samples),
                                     title = "PCA Plot of Log2 Raw Counts")
pcas_deseq2_before$pca2d

# exit_gracefully()
# print(pca_plot)
```

### After normalization

```{r plot_pca_plotly_deseq2_after}
#| code-fold: true
#| code-summary: "🧬 PCA Plot with Group Colors and Variance Labels"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false

if (isTRUE(params$usebatch)) {
  pcas_deseq2_after <- make_pca_plots(normalized_counts_deseq2_for_plots,as.data.frame(samples),
                                       title = "PCA Plot of VST Normalized Counts (Batch corrected)")
} else {
  pcas_deseq2_after <- make_pca_plots(normalized_counts_deseq2_for_plots,as.data.frame(samples),
                                       title = "PCA Plot of VST Normalized Counts")
}

pcas_deseq2_after$pca2d

```

:::

## 3-D PCA plot

::: {.panel-tabset}

### Before normalization

```{r plot_pca_3d_plotly_deseq2_before}
#| code-fold: true
#| code-summary: "🧬 3D PCA Plot"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false


pcas_deseq2_before$pca3d
# exit_gracefully()
# print(pca_3d_plot)
```

### After normalization

```{r plot_pca_3d_plotly_deseq2_after}
#| code-fold: true
#| code-summary: "🧬 3D PCA Plot"
#| fig.width: 8
#| fig.height: 6
#| out.width: "100%"
#| out.height: "100%"
#| fig.align: "center"
#| fig.show: "hold"
#| warning: false


pcas_deseq2_after$pca3d
```

:::


